---
title: "Fitting traits to data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting traits to data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 120, fig.align = "center")
```

```{r setup, message = FALSE}
library(phylloptim)
```

`vignette("phylloptim")` shows the model solving and differentiating. This one
uses both to recover traits from data, and is honest about the three things that
make it harder than it looks.

## Simulated data, so the answer is known

Recovery error is the only fit diagnostic that distinguishes "the optimiser
converged" from "the model works". Real data cannot supply it, so start here.

```{r}
set.seed(sum(utf8ToInt("phylloptim-fitting")))

truth <- leaf_traits(vcmax_25 = 85, jmax_25 = 140, cost_scale_TF24 = 40)

drivers <- expand.grid(
  psi_soil = seq(0.4, 3.2, length.out = 12),
  PPFD = c(400, 900, 1500),
  atm_vpd = c(1.0, 2.5)
)

obs <- leaf_solve(psi_soil = drivers$psi_soil, PPFD = drivers$PPFD,
                  atm_vpd = drivers$atm_vpd, traits = truth)

sigma <- c(A = 0.4, gc = 0.0015, psi_stem = 0.08)
obs$A_obs <- obs$A + rnorm(nrow(obs), 0, sigma[["A"]])
obs$gc_obs <- obs$gc + rnorm(nrow(obs), 0, sigma[["gc"]])
obs$psi_obs <- obs$psi_stem + rnorm(nrow(obs), 0, sigma[["psi_stem"]])

nrow(obs)
```

Three responses, because they constrain different parts of the model: `psi_stem`
pins the hydraulic path, `A` and `gc` pin the carbon side. Fitting `A` alone
leaves the two trading off against each other.

## The objective, and the loop that makes it affordable

```{r}
FIT <- c("vcmax_25", "jmax_25", "cost_scale_TF24")

predict_at <- function(theta, l) {
  set_traits(l, do.call(leaf_traits, as.list(exp(theta))))
  out <- vapply(seq_len(nrow(obs)), function(i) {
    set_drivers(l, psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                atm_vpd = obs$atm_vpd[i])
    l$find_root_collar_psi()
    c(l$assim_colimited_, l$stom_cond_CO2_, l$opt_psi_stem_)
  }, numeric(3))
  list(A = out[1, ], gc = out[2, ], psi = out[3, ])
}

ssq <- function(theta, l) {
  p <- predict_at(theta, l)
  sum(((obs$A_obs - p$A) / sigma[["A"]])^2) +
    sum(((obs$gc_obs - p$gc) / sigma[["gc"]])^2) +
    sum(((obs$psi_obs - p$psi) / sigma[["psi_stem"]])^2)
}
```

⚠️ **Build the model once, outside the loop.** `set_traits()` replaces the traits
on an existing object; constructing a fresh one costs ~204 µs against ~4 µs to
re-trait, which is 73 solves thrown away per evaluation. In a fit that dominates
everything else.

⚠️ **And do not write `l$vcmax_25 <- x` instead.** It compiles and is wrong: the
derived `vcmax_`, `jmax_` and `R_d_` live in a temperature cache keyed on
`(leaf_temp, atm_o2_kpa)` and *nothing else*, so changing a trait and re-setting
the drivers takes a cache **hit** and never recomputes them. A sweep written that
way runs at the first `vcmax` it ever saw and reports plausible numbers
throughout. `set_traits()` refreshes all of it.

```{r}
l <- leaf_model(truth)
start <- log(c(vcmax_25 = 60, jmax_25 = 100, cost_scale_TF24 = 15))
fit <- optim(start, ssq, l = l, method = "Nelder-Mead",
             control = list(maxit = 2000, reltol = 1e-10))

data.frame(
  parameter = FIT,
  truth = unlist(truth[FIT]),
  recovered = round(exp(fit$par), 2),
  `pct error` = round(100 * (exp(fit$par) - unlist(truth[FIT])) /
                        unlist(truth[FIT]), 2),
  check.names = FALSE
)
```

## Using the gradients

`leaf_gradient()` gives exact derivatives of each output, so the objective's
gradient follows by the chain rule — no differencing of the solve.

```{r}
ssq_grad <- function(theta, l) {
  tr <- do.call(leaf_traits, as.list(exp(theta)))
  g <- vapply(seq_len(nrow(obs)), function(i) {
    gi <- leaf_gradient(psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                        atm_vpd = obs$atm_vpd[i], traits = tr, pars = FIT)
    v <- gi$value
    r <- c((v[["A"]] - obs$A_obs[i]) / sigma[["A"]]^2,
           (v[["gc"]] - obs$gc_obs[i]) / sigma[["gc"]]^2,
           (v[["psi_stem"]] - obs$psi_obs[i]) / sigma[["psi_stem"]]^2)
    2 * (gi$gradient[FIT, "A"] * r[1] +
           gi$gradient[FIT, "gc"] * r[2] +
           gi$gradient[FIT, "psi_stem"] * r[3])
  }, numeric(length(FIT)))
  rowSums(g) * exp(theta)   # chain rule for the log parameterisation
}
```

Note `pars = FIT`. Requesting all sixteen parameters here would compute thirteen
gradients per observation that the optimiser never reads.

⚠️ **Bound the parameters, or the first step overflows.** The objective is scaled
by `1/sigma^2` and `sigma_gc` is 0.0015, so the gradient has magnitude ~2e4 in log
units. Plain `method = "BFGS"` starts from an identity Hessian approximation, which
makes its first trial point `theta - g` — and `exp()` of that is `Inf`, so the
objective errors out before the line search gets a chance to shrink the step. This
is not a pathology of the model; it is what an unscaled least-squares objective
does to any quasi-Newton method started cold. Bounds are the one-line fix, and are
what you would want on a real fit anyway.

```{r}
lower <- log(c(vcmax_25 = 10, jmax_25 = 20, cost_scale_TF24 = 2))
upper <- log(c(vcmax_25 = 400, jmax_25 = 700, cost_scale_TF24 = 400))

fit_g <- optim(start, ssq, gr = ssq_grad, l = l, method = "L-BFGS-B",
               lower = lower, upper = upper, control = list(maxit = 500))

fit_n <- optim(start, ssq, l = l, method = "L-BFGS-B",
               lower = lower, upper = upper, control = list(maxit = 500))

data.frame(
  method = c("Nelder-Mead (no gradient)", "L-BFGS-B (numerical gradient)",
             "L-BFGS-B (exact gradient)"),
  objective = round(c(fit$value, fit_n$value, fit_g$value), 2),
  `objective evals` = c(fit$counts[["function"]], fit_n$counts[["function"]],
                        fit_g$counts[["function"]]),
  `gradient evals` = c(0, fit_n$counts[["gradient"]], fit_g$counts[["gradient"]]),
  `honest total` = c(fit$counts[["function"]],
                     fit_n$counts[["function"]] +
                       2 * length(FIT) * fit_n$counts[["gradient"]],
                     fit_g$counts[["function"]]),
  check.names = FALSE
)
```

All three find the same optimum. The exact gradient cuts the solve count roughly
five-fold against Nelder-Mead — and **that is the whole of its advantage here.**

⚠️ **Counting evaluations fairly is harder than it looks.** `optim` reports
function and gradient calls separately, and when it differences a gradient
*numerically* it does not count those calls at all. BFGS and L-BFGS-B use central
differences, so each numerical gradient silently costs `2 × npar` extra objective
evaluations — the `honest total` column above adds them back. Comparing a
derivative-free method's reported count against a gradient method's reported count
without that correction flatters the gradient method by a factor of `2 × npar`.

## ⚠️ The exact gradient is currently *slower* than differencing it

Evaluation counts are not wall clock, and here the two disagree. Measure both arms
in one process, interleaved:

```{r}
fd_gradient <- function() {
  for (j in seq_along(start)) {
    a <- start; b <- start; a[j] <- a[j] + 1e-5; b[j] <- b[j] - 1e-5
    ssq(a, l); ssq(b, l)
  }
}
construct_only <- function() for (i in seq_len(nrow(obs))) leaf_model(truth)

# Warm up all three arms first, then interleave. A cold leaf_model() measured
# three times its steady-state cost, which is enough to turn "a third" into
# "nine tenths" and point the conclusion at the wrong fix.
for (k in 1:3) { ssq_grad(start, l); fd_gradient(); construct_only() }

tm <- function(f) {
  t0 <- Sys.time(); f(); 1000 * as.numeric(difftime(Sys.time(), t0, units = "secs"))
}
reps <- replicate(5, c(tm(function() ssq_grad(start, l)),
                       tm(fd_gradient),
                       tm(construct_only)))
data.frame(
  route = c("exact gradient (72 x leaf_gradient)",
            "central differences (2 x npar objective evals)",
            "72 x leaf_model() alone"),
  ms = round(apply(reps, 1, min), 1)
)
```

The exact gradient loses on wall clock while winning on evaluation count, and the
third row is **about a third** of the reason. `leaf_gradient()` constructs a fresh
model on every call — there is no argument for handing it one you already have — so
a gradient over `n` observations pays `n` constructions.

The other two thirds is the R call boundary itself. The composite replaces one
solve with a handful of much cheaper evaluations, but each of those is an R call
into C++, and an R call costs more than the C++ work it wraps. The theorem does
less arithmetic and more crossing.

Three things follow:

- **Do not choose between these on the counts alone.** Time them on your own design.
  Where the objective is expensive relative to a model construction — many more
  observations, or a supply path with more soil layers — the ranking flips back.
- **Model reuse is worth a third of it**, and is tracked as
  [phylloptim#52](https://github.com/traitecoevo/phylloptim/issues/52): let
  `leaf_gradient()` accept an existing model the way `set_traits()` does. On its own
  it would not make the exact gradient the faster arm here.
- **From C++ neither overhead exists.** `vignette("cpp-interface")` has the same
  comparison with no R in the way. Moving the composition into C++ is stage 2 of
  PLAN item 11, and it is what would actually reverse this result.

## ⚠️ Where a gradient-based fit goes wrong

The derivation behind `leaf_gradient()` assumes the optimum is interior. In dry
soil it often is not, and the boundary is a genuine kink in trait space rather
than an artefact.

```{r}
status <- vapply(seq_len(nrow(obs)), function(i) {
  leaf_gradient(psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                atm_vpd = obs$atm_vpd[i], traits = truth,
                pars = "vcmax_25")$status
}, character(1))
table(status)
```

**A third of the design is pinned**, and the fit above still recovered the traits
to a few percent — because `leaf_gradient()` detects those points and differences
the solve there instead of applying a formula that would be wrong by orders of
magnitude. That is the fallback doing its job on a perfectly ordinary drought
design, not an exotic case.

Check this *before* trusting a fit rather than after. `status` is reported
precisely so it can be checked, and a naive finite difference cannot tell you
which points are which. Where the pinned fraction is large, expect a
gradient-based optimiser to behave less well than it does here.

Because a fit that straddles the boundary has three problems:

- **the objective is not differentiable there**, so a gradient method is
  optimising through a corner and may stall or oscillate;
- `leaf_gradient()` falls back to differencing the solve at pinned points, which
  is correct but noisier;
- ⚠️ **`psi_crit` is special.** It does not appear in the profit function at all
  — it only bounds the feasible range — so the composite returns **exactly zero**
  for it. At an interior optimum that is right. At a dry-pinned optimum it is
  binding and the true derivative is not zero, so a silent zero would tell an
  optimiser the parameter does nothing. The fallback exists for exactly this and
  is not a degraded path.

## Fitting to real data

The same machinery, with two additions worth planning for.

**Not every parameter is identified.** Parameters that enter the model only in
combination cannot be separated by data that sees only the combination — fitting
them independently produces respectable-looking estimates that mean nothing.
Check with a profile, or a Hessian at the optimum, before reporting anything
per-parameter.

**Some traits are measured.** `P50` and `P88` determine the Weibull
`stem_b`/`stem_c` analytically, so a species with a published vulnerability curve
needs no hydraulic calibration at all:

```{r}
weibull_from_p50_p88 <- function(P50, P88) {
  c_ <- log(log(1 / 0.12) / log(2)) / log(P88 / P50)
  c(stem_b = P50 / log(2)^(1 / c_), stem_c = c_)
}
w <- weibull_from_p50_p88(P50 = 6.88, P88 = 9.43)
round(w, 4)

lw <- leaf_model(leaf_traits(stem_b = w[["stem_b"]], stem_c = w[["stem_c"]]))
round(c(PLC_at_P50 = 1 - lw$proportion_of_conductivity(6.88),
        PLC_at_P88 = 1 - lw$proportion_of_conductivity(9.43)), 4)
```

Exactly 0.50 and 0.88, as they must be. Fitting what is already measured wastes
the data's ability to constrain what is not.
