---
title: "Fitting traits to data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting traits to data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 120, fig.align = "center")
```

```{r setup, message = FALSE}
library(phylloptim)
```

`vignette("phylloptim")` shows the model solving and differentiating. This one
uses both to recover traits from data, and is honest about the three things that
make it harder than it looks.

## Simulated data, so the answer is known

Recovery error is the only fit diagnostic that distinguishes "the optimiser
converged" from "the model works". Real data cannot supply it, so start here.

```{r}
set.seed(sum(utf8ToInt("phylloptim-fitting")))

truth <- leaf_traits(vcmax_25 = 85, jmax_25 = 140, cost_scale_TF24 = 40)

drivers <- expand.grid(
  psi_soil = seq(0.4, 3.2, length.out = 12),
  PPFD = c(400, 900, 1500),
  atm_vpd = c(1.0, 2.5)
)

obs <- leaf_solve(psi_soil = drivers$psi_soil, PPFD = drivers$PPFD,
                  atm_vpd = drivers$atm_vpd, traits = truth)

sigma <- c(A = 0.4, gc = 0.0015, psi_stem = 0.08)
obs$A_obs <- obs$A + rnorm(nrow(obs), 0, sigma[["A"]])
obs$gc_obs <- obs$gc + rnorm(nrow(obs), 0, sigma[["gc"]])
obs$psi_obs <- obs$psi_stem + rnorm(nrow(obs), 0, sigma[["psi_stem"]])

nrow(obs)
```

Three responses, because they constrain different parts of the model: `psi_stem`
pins the hydraulic path, `A` and `gc` pin the carbon side. Fitting `A` alone
leaves the two trading off against each other.

## The objective, and the loop that makes it affordable

```{r}
FIT <- c("vcmax_25", "jmax_25", "cost_scale_TF24")

predict_at <- function(theta, l) {
  set_traits(l, do.call(leaf_traits, as.list(exp(theta))))
  out <- vapply(seq_len(nrow(obs)), function(i) {
    set_drivers(l, psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                atm_vpd = obs$atm_vpd[i])
    l$find_root_collar_psi()
    c(l$assim_colimited_, l$stom_cond_CO2_, l$opt_psi_stem_)
  }, numeric(3))
  list(A = out[1, ], gc = out[2, ], psi = out[3, ])
}

ssq <- function(theta, l) {
  p <- predict_at(theta, l)
  sum(((obs$A_obs - p$A) / sigma[["A"]])^2) +
    sum(((obs$gc_obs - p$gc) / sigma[["gc"]])^2) +
    sum(((obs$psi_obs - p$psi) / sigma[["psi_stem"]])^2)
}
```

⚠️ **Build the model once, outside the loop.** `set_traits()` replaces the traits
on an existing object; constructing a fresh one costs ~204 µs against ~4 µs to
re-trait, which is 73 solves thrown away per evaluation. In a fit that dominates
everything else.

⚠️ **And do not write `l$vcmax_25 <- x` instead.** It compiles and is wrong: the
derived `vcmax_`, `jmax_` and `R_d_` live in a temperature cache keyed on
`(leaf_temp, atm_o2_kpa)` and *nothing else*, so changing a trait and re-setting
the drivers takes a cache **hit** and never recomputes them. A sweep written that
way runs at the first `vcmax` it ever saw and reports plausible numbers
throughout. `set_traits()` refreshes all of it.

```{r}
l <- leaf_model(truth)
start <- log(c(vcmax_25 = 60, jmax_25 = 100, cost_scale_TF24 = 15))
fit <- optim(start, ssq, l = l, method = "Nelder-Mead",
             control = list(maxit = 2000, reltol = 1e-10))

data.frame(
  parameter = FIT,
  truth = unlist(truth[FIT]),
  recovered = round(exp(fit$par), 2),
  `pct error` = round(100 * (exp(fit$par) - unlist(truth[FIT])) /
                        unlist(truth[FIT]), 2),
  check.names = FALSE
)
```

## Using the gradients

`leaf_gradient()` gives exact derivatives of each output, so the objective's
gradient follows by the chain rule — no differencing of the solve.

```{r}
ssq_grad <- function(theta, l) {
  tr <- do.call(leaf_traits, as.list(exp(theta)))
  g <- vapply(seq_len(nrow(obs)), function(i) {
    gi <- leaf_gradient(psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                        atm_vpd = obs$atm_vpd[i], traits = tr, pars = FIT)
    v <- gi$value
    r <- c((v[["A"]] - obs$A_obs[i]) / sigma[["A"]]^2,
           (v[["gc"]] - obs$gc_obs[i]) / sigma[["gc"]]^2,
           (v[["psi_stem"]] - obs$psi_obs[i]) / sigma[["psi_stem"]]^2)
    2 * (gi$gradient[FIT, "A"] * r[1] +
           gi$gradient[FIT, "gc"] * r[2] +
           gi$gradient[FIT, "psi_stem"] * r[3])
  }, numeric(length(FIT)))
  rowSums(g) * exp(theta)   # chain rule for the log parameterisation
}
```

Note `pars = FIT`. Requesting all fourteen parameters here would compute eleven
gradients per observation that the optimiser never reads.

⚠️ **Bound the parameters, or the first step overflows.** The objective is scaled
by `1/sigma^2` and `sigma_gc` is 0.0015, so the gradient has magnitude ~2e4 in log
units. Plain `method = "BFGS"` starts from an identity Hessian approximation, which
makes its first trial point `theta - g` — and `exp()` of that is `Inf`, so the
objective errors out before the line search gets a chance to shrink the step. This
is not a pathology of the model; it is what an unscaled least-squares objective
does to any quasi-Newton method started cold. Bounds are the one-line fix, and are
what you would want on a real fit anyway.

```{r}
lower <- log(c(vcmax_25 = 10, jmax_25 = 20, cost_scale_TF24 = 2))
upper <- log(c(vcmax_25 = 400, jmax_25 = 700, cost_scale_TF24 = 400))

fit_g <- optim(start, ssq, gr = ssq_grad, l = l, method = "L-BFGS-B",
               lower = lower, upper = upper, control = list(maxit = 500))

fit_n <- optim(start, ssq, l = l, method = "L-BFGS-B",
               lower = lower, upper = upper, control = list(maxit = 500))

data.frame(
  method = c("Nelder-Mead (no gradient)", "L-BFGS-B (numerical gradient)",
             "L-BFGS-B (exact gradient)"),
  objective = round(c(fit$value, fit_n$value, fit_g$value), 2),
  `objective evals` = c(fit$counts[["function"]], fit_n$counts[["function"]],
                        fit_g$counts[["function"]]),
  `gradient evals` = c(0, fit_n$counts[["gradient"]], fit_g$counts[["gradient"]]),
  `honest total` = c(fit$counts[["function"]],
                     fit_n$counts[["function"]] +
                       2 * length(FIT) * fit_n$counts[["gradient"]],
                     fit_g$counts[["function"]]),
  check.names = FALSE
)
```

All three find the same optimum. The exact gradient cuts the solve count roughly
five-fold against Nelder-Mead — and **that is the whole of its advantage here.**

⚠️ **Counting evaluations fairly is harder than it looks.** `optim` reports
function and gradient calls separately, and when it differences a gradient
*numerically* it does not count those calls at all. BFGS and L-BFGS-B use central
differences, so each numerical gradient silently costs `2 × npar` extra objective
evaluations — the `honest total` column above adds them back. Comparing a
derivative-free method's reported count against a gradient method's reported count
without that correction flatters the gradient method by a factor of `2 × npar`.

## ⚠️ Faster or slower depends on the parameter count — and on *which* parameter count

Evaluation counts are not wall clock, and here the two disagree: at three
parameters the exact gradient **loses**. That is a real result, and it is also the
most misread one in this package, because it does not generalise the way it looks
like it should. Both arms scale with the number of parameters, and they scale with
**different** parameter counts.

Measure both arms in one process, interleaved, over a growing request:

```{r}
#| label: scaling
tm <- function(f) {
  t0 <- Sys.time(); f(); 1000 * as.numeric(difftime(Sys.time(), t0, units = "secs"))
}

CAND <- c("vcmax_25", "jmax_25", "cost_scale_TF24", "beta2", "a",
          "curv_fact_elec_trans", "curv_fact_colim", "psi_crit",
          "root_psi_crit", "stem_b", "stem_c", "root_b", "root_c")

exact_n <- function(pars) {
  vapply(seq_len(nrow(obs)), function(i) {
    leaf_gradient(psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                  atm_vpd = obs$atm_vpd[i], traits = truth,
                  pars = pars)$gradient[pars, "A"]
  }, numeric(length(pars)))
}
ssq_at <- function(theta, l, nm) {
  tr <- truth; tr[nm] <- as.list(exp(theta))
  set_traits(l, do.call(leaf_traits, tr))
  out <- vapply(seq_len(nrow(obs)), function(i) {
    set_drivers(l, psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                atm_vpd = obs$atm_vpd[i])
    l$find_root_collar_psi()
    c(l$assim_colimited_, l$stom_cond_CO2_, l$opt_psi_stem_)
  }, numeric(3))
  sum(((obs$A_obs - out[1, ]) / sigma[["A"]])^2) +
    sum(((obs$gc_obs - out[2, ]) / sigma[["gc"]])^2) +
    sum(((obs$psi_obs - out[3, ]) / sigma[["psi_stem"]])^2)
}
fd_n <- function(pars, l) {
  th <- log(unlist(truth[pars]))
  for (j in seq_along(pars)) {
    a <- th; b <- th; a[j] <- a[j] + 1e-5; b[j] <- b[j] - 1e-5
    ssq_at(a, l, pars); ssq_at(b, l, pars)
  }
}

for (k in 1:3) { invisible(exact_n(CAND[1:2])); fd_n(CAND[1:2], l) }  # warm both

scaling <- do.call(rbind, lapply(c(1, 3, 5, 8, 13), function(n) {
  p <- CAND[seq_len(n)]
  r <- replicate(3, c(tm(function() exact_n(p)), tm(function() fd_n(p, l))))
  data.frame(n_par = n, exact_ms = round(min(r[1, ]), 1),
             fd_ms = round(min(r[2, ]), 1))
}))
scaling$ratio <- round(scaling$fd_ms / scaling$exact_ms, 2)
scaling
```

```{r}
#| label: scaling-slopes
rbind(
  exact = round(coef(lm(exact_ms ~ n_par, scaling)), 2),
  fd = round(coef(lm(fd_ms ~ n_par, scaling)), 2)
)
```

Both are linear, and the two coefficients are the whole story:

- **the finite difference is pure slope** — its intercept is ~0 and its slope is
  `2 ×` one objective evaluation, because that is literally what it does;
- **the exact gradient has a large intercept and a slope too.** The intercept is
  the per-observation `Leaf` construction and solve, paid once whatever you ask
  for. The slope is `2` evaluations per parameter per observation.

Here the exact route's *slope* is the larger of the two, so **growing the parameter
count in this design never makes it win** — the ratio improves but does not reach 1.
Each extra parameter costs it two more R calls per observation, and from R a call
costs more than the C++ work it wraps.

### ⚠️ So what reverses it is not more parameters, but more *fitted* parameters per *model* parameter

The two counts are the same thing only because this vignette fits traits directly.
They come apart as soon as a parameterisation sits between the fit and the model:

$$T_\text{FD} = 2\,P_\text{fit}\,T_\text{obj}
\qquad
T_\text{exact} = N\,(t_\text{construct} + t_\text{solve}) + 2\,P_\text{model}\,N\,t_\text{eval}$$

**`leaf_gradient()` is asked for `P_model` parameters; `optim` differences
`P_fit` of them.** A hierarchical model is the standard case where those differ: a
non-centred population mean, SD and per-species deviation are three fitted
parameters that reach the leaf as *one*.

The companion study
[`leaf-calibration`](https://github.com/traitecoevo/leaf_calibration_test) is that
case, and it is worth quoting because it comes out the other way round. 1,327
observations, 16 species, **40 fitted parameters that map onto 4 model
parameters**:

| | |
|---|---|
| exact gradient, 40 fitted parameters | **0.68 s** |
| exact gradient, 57 fitted parameters | **0.67 s** — unchanged, because `P_model` is still 4 |
| central differences, 40 fitted parameters | **2.12 s** |
| the resulting fit | same optimum, **3.4× less wall clock** |

⚠️ **The cost model is not fitted to that study — it PREDICTS it.** Take the two
coefficients measured above, divide by this vignette's 72 observations to get
per-observation costs, and substitute that study's design:

```{r}
#| label: predict-other-study
co <- coef(lm(exact_ms ~ n_par, scaling)) / nrow(obs)   # ms per observation
LC <- list(N = 1327, P_model = 4, T_obj_ms = 26,        # measured there
           exact_ms = 679, break_even = 13.1)

predicted <- LC$N * (co[[1]] + co[[2]] * LC$P_model)
data.frame(
  quantity = c("exact gradient (ms)", "break-even fitted parameters"),
  predicted = round(c(predicted, predicted / (2 * LC$T_obj_ms)), 1),
  measured_there = c(LC$exact_ms, LC$break_even)
)
```

Two designs differing 18-fold in size, one set of coefficients, agreeing to a few
percent. That is the evidence that the scaling law is the mechanism and not a
coincidence of either design.

**So the rule to take away is not "AD is slower here".** It is:

> Count how many *distinct model parameters* your gradient needs, not how many
> parameters your optimiser is moving. Differencing costs the latter; the composite
> costs the former.

If those numbers are equal, difference the objective. If your parameterisation
makes the first much smaller than the second — pooling, hierarchy, any shared or
derived parameter — use the gradient.

### Where the intercept goes

The remaining measurement, since the intercept is the exact route's whole
disadvantage in this design:

```{r}
fd_gradient <- function() {
  for (j in seq_along(start)) {
    a <- start; b <- start; a[j] <- a[j] + 1e-5; b[j] <- b[j] - 1e-5
    ssq(a, l); ssq(b, l)
  }
}
construct_only <- function() for (i in seq_len(nrow(obs))) leaf_model(truth)

# Warm up all three arms first, then interleave. A cold leaf_model() measured
# three times its steady-state cost, which is enough to turn "a third" into
# "nine tenths" and point the conclusion at the wrong fix.
for (k in 1:3) { ssq_grad(start, l); fd_gradient(); construct_only() }

reps <- replicate(5, c(tm(function() ssq_grad(start, l)),
                       tm(fd_gradient),
                       tm(construct_only)))
data.frame(
  route = c("exact gradient (72 x leaf_gradient)",
            "central differences (2 x npar objective evals)",
            "72 x leaf_model() alone"),
  ms = round(apply(reps, 1, min), 1)
)
```

The third row is most of the intercept. `leaf_gradient()` constructs a fresh model
on every call — there is no argument for handing it one you already have — so a
gradient over `n` observations pays `n` constructions.

The rest is the R call boundary itself. The composite replaces one solve with a
handful of much cheaper evaluations, but each of those is an R call into C++, and an
R call costs more than the C++ work it wraps. The theorem does less arithmetic and
more crossing — which is also why the exact route has a *slope* in `P_model` at all,
rather than the near-free one the C++ benchmarks show.

Four things follow:

- **Do not choose between these on the evaluation counts alone**, and do not choose
  on this vignette's verdict either. Put your own `N`, `P_fit` and `P_model` into the
  cost model above, or time it on your own design.
- **`P_fit` = `P_model` is the worst case for the composite**, and it is the case a
  vignette naturally demonstrates because fitting traits directly is the simplest
  thing to show. Real fits are usually not that case.
- **Model reuse was most of the intercept, and it is fixed**
  ([phylloptim#52](https://github.com/traitecoevo/phylloptim/issues/52)):
  `leaf_gradient(x = )` takes a model you already have. In `leaf-calibration` that
  was 29% of the exact gradient. It did not make the composite the faster arm
  *here*, because here the problem is the slope.
- **And the slope is gone too, which supersedes this whole comparison.**
  `leaf_gradient_batch()` composes the theorem in C++ and loops over observations
  there, so the boundary is crossed once per call instead of 112 times per
  observation: **363 → 10.6 µs per observation** at four fitted parameters, 22×.

⚠️ **So do not read the table above as the state of the art — read it as what the R
boundary costs.** It is kept because the cost *model* is still the thing to reason
with, and because `P_fit` versus `P_model` still decides which arm wins. What has
changed is only the constant in front of the exact arm. The same fit through
`leaf_batch()` + `leaf_gradient_batch()`:

```r
b <- leaf_batch(psi_soil = obs$psi_soil, PPFD = obs$PPFD)   # once, not per draw
g <- leaf_gradient_batch(b, traits, pars = FIT)
```

with the likelihood and your parameterisation Jacobian still in R, vectorised over
observations — which is where they belong, and is why the `P_fit > P_model`
structure is what the exact route exploits.

## ⚠️ Where a gradient-based fit goes wrong

The derivation behind `leaf_gradient()` assumes the optimum is interior. In dry
soil it often is not, and the boundary is a genuine kink in trait space rather
than an artefact.

```{r}
status <- vapply(seq_len(nrow(obs)), function(i) {
  leaf_gradient(psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                atm_vpd = obs$atm_vpd[i], traits = truth,
                pars = "vcmax_25")$status
}, character(1))
table(status)
```

**A third of the design is pinned**, and the fit above still recovered the traits
to a few percent — because `leaf_gradient()` detects those points and differences
the solve there instead of applying a formula that would be wrong by orders of
magnitude. That is the fallback doing its job on a perfectly ordinary drought
design, not an exotic case.

Check this *before* trusting a fit rather than after. `status` is reported
precisely so it can be checked, and a naive finite difference cannot tell you
which points are which. Where the pinned fraction is large, expect a
gradient-based optimiser to behave less well than it does here.

Because a fit that straddles the boundary has three problems:

- **the objective is not differentiable there**, so a gradient method is
  optimising through a corner and may stall or oscillate;
- `leaf_gradient()` falls back to differencing the solve at pinned points, which
  is correct but noisier;
- ⚠️ **`psi_crit` is special.** It does not appear in the profit function at all
  — it only bounds the feasible range — so the composite returns **exactly zero**
  for it. At an interior optimum that is right. At a dry-pinned optimum it is
  binding and the true derivative is not zero, so a silent zero would tell an
  optimiser the parameter does nothing. The fallback exists for exactly this and
  is not a degraded path.

## Fitting to real data

The same machinery, with two additions worth planning for.

**Not every parameter is identified.** Parameters that enter the model only in
combination cannot be separated by data that sees only the combination — fitting
them independently produces respectable-looking estimates that mean nothing.
Check with a profile, or a Hessian at the optimum, before reporting anything
per-parameter.

**Some traits are measured.** `P50` and `P88` determine the Weibull
`stem_b`/`stem_c` analytically, so a species with a published vulnerability curve
needs no hydraulic calibration at all:

```{r}
weibull_from_p50_p88 <- function(P50, P88) {
  c_ <- log(log(1 / 0.12) / log(2)) / log(P88 / P50)
  c(stem_b = P50 / log(2)^(1 / c_), stem_c = c_)
}
w <- weibull_from_p50_p88(P50 = 6.88, P88 = 9.43)
round(w, 4)

lw <- leaf_model(leaf_traits(stem_b = w[["stem_b"]], stem_c = w[["stem_c"]]))
round(c(PLC_at_P50 = 1 - lw$proportion_of_conductivity(6.88),
        PLC_at_P88 = 1 - lw$proportion_of_conductivity(9.43)), 4)
```

Exactly 0.50 and 0.88, as they must be. Fitting what is already measured wastes
the data's ability to constrain what is not.
