---
title: "Leaf gas exchange from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Leaf gas exchange from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, dpi = 120)
```

```{r setup}
library(leaf)
```

## What this model does differently

Most leaf gas-exchange models get stomatal conductance from an empirical
function of assimilation, VPD and CO₂ — Ball-Berry, Medlyn, and their
descendants. This one does not have such a function. It builds an explicit
hydraulic path from the soil, through the roots and the root collar, up the
stem to the leaf, and then asks: **which stem water potential maximises profit?**
Profit is assimilation minus a hydraulic cost that rises as the xylem loses
conductivity. Conductance is whatever falls out of the answer.

The consequence worth caring about is that stomatal closure under drought is not
fitted, it is *derived* — from two Weibull vulnerability curves, the soil
potential, and the plant's own conductance.

## One solve

The object is stateful, which is unusual for R but is what makes it fast enough
to sit inside a demographic model that calls it millions of times: you set the
drivers, solve, and read the operating point off the object.

⚠️ **Nineteen positional trait arguments is not a reasonable thing to type at a
console**, and it is not the intended long-term interface — it is the C++
constructor, exposed as-is so that the R layer could be proved correct against
the C++ regression baseline before being redesigned. A named, defaulted
interface is the next piece of work.

```{r}
saligna <- function() {
  Leaf(vcmax_25 = 96, stem_c = 2.680147, stem_b = 3.898245,
       psi_crit = 5.870283, root_c = 2.680147, root_b = 3.898245,
       root_psi_crit = 5.870283, beta2 = 1.5, jmax_25 = 157.44, a = 0.30,
       curv_fact_elec_trans = 0.7, curv_fact_colim = 0.99,
       GSS_tol_abs = 1e-3, vulnerability_curve_ncontrol = 100,
       ci_abs_tol = 1e-3, ci_niter = 1000, cost_scale_TF24 = 7.5,
       beta_R_H = 3.4e2, beta_R_V = 9.4e3)
}

l <- saligna()
l$set_physiology(
  root_carbon_per_leaf_area = 20,   # kg C per m2 LEAF, not absolute carbon
  PPFD = 900,                       # umol m-2 s-1
  psi_soil = 2.0,                   # MPa, POSITIVE magnitude -- see below
  soil_depth = 1.0,                 # m
  leaf_specific_conductance_max = 3.14e-5,
  atm_vpd = 2.0, ca = 40.0, leaf_temp = 25,
  atm_o2_kpa = 21, atm_kpa = 101.3
)
l$find_root_collar_psi()

c(psi_stem = l$opt_psi_stem_,   # leaf water potential at the optimum, MPa
  collar   = l$opt_root_psi_,   # root-collar potential, MPa
  A        = l$assim_colimited_,
  E        = l$transpiration_,
  gc       = l$stom_cond_CO2_,
  profit   = l$profit_)
```

Two derived quantities that are the reason for comparing this model against
others. `lambda` is the marginal cost of water, dA/dE, at the operating point;
`g1_eff` is the Medlyn `g1` that an empirical model would have to be given to
reproduce this leaf's behaviour here.

```{r}
c(lambda = l$lambda, g1_eff = l$g1_eff)
```

The point of `g1_eff` is that it is not a constant. An empirical model treats
`g1` as a trait; here it is an *output*, and watching it move with soil moisture
is watching the empirical model's central assumption fail.

## Water potentials are positive magnitudes

There is one representation throughout: every ψ is a positive number of MPa,
including `psi_soil`, `psi_crit` and the collar potential. Where an equation
needs one potential to oppose another, the minus sign is in the equation.

This is asserted rather than documented, which matters if you are porting an
older script that passed signed potentials:

```{r, error = TRUE}
l$set_physiology(
  root_carbon_per_leaf_area = 20, PPFD = 900,
  psi_soil = -2.0, soil_depth = 1.0,
  leaf_specific_conductance_max = 3.14e-5,
  atm_vpd = 2.0, ca = 40.0, leaf_temp = 25,
  atm_o2_kpa = 21, atm_kpa = 101.3
)
```

An error, not a quiet `abs()`. A model that accepted both conventions would let
that script return a plausible wrong number instead.

## A drought response, derived rather than fitted

Sweep soil potential from wet to very dry and watch the leaf shut down. Nothing
in the code below tells it when to close.

```{r}
solve_at <- function(psi_soil, vpd = 2.0, ppfd = 900) {
  l <- saligna()
  l$set_physiology(root_carbon_per_leaf_area = 20, PPFD = ppfd,
                   psi_soil = psi_soil, soil_depth = 1.0,
                   leaf_specific_conductance_max = 3.14e-5,
                   atm_vpd = vpd, ca = 40.0, leaf_temp = 25,
                   atm_o2_kpa = 21, atm_kpa = 101.3)
  l$find_root_collar_psi()
  data.frame(psi_soil = psi_soil, vpd = vpd,
             A = l$assim_colimited_, E = l$transpiration_,
             gc = l$stom_cond_CO2_, psi_stem = l$opt_psi_stem_,
             g1_eff = l$g1_eff)
}

drought <- do.call(rbind, lapply(seq(0.2, 5.5, length.out = 40), solve_at))

op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2, 1))
plot(A ~ psi_soil, drought, type = "l", lwd = 2,
     xlab = "soil water potential (MPa)",
     ylab = expression(A~(mu*mol~m^-2~s^-1)),
     main = "assimilation")
plot(gc ~ psi_soil, drought, type = "l", lwd = 2, col = "steelblue",
     xlab = "soil water potential (MPa)",
     ylab = expression(g[c]~(mol~m^-2~s^-1)),
     main = "stomatal conductance")
par(op)
```

Note the point where both hit exactly zero. That is not a floor imposed by a
minimum-conductance parameter — it is the solve reporting that no operating
point both moves water and stays inside the root's hydraulic limit, so the
answer is shut-down.

## `g1_eff` is not a constant

The same sweep, read as the empirical parameter it would take to mimic this
leaf:

```{r}
plot(g1_eff ~ psi_soil, drought[is.finite(drought$g1_eff) & drought$gc > 0, ],
     type = "l", lwd = 2, col = "darkred",
     xlab = "soil water potential (MPa)",
     ylab = expression(g[1]~effective~(kPa^0.5)),
     main = "the Medlyn g1 this leaf implies")
```

A Medlyn model fitted at one soil moisture and applied at another is using the
wrong `g1`, by a factor this curve measures. That is the argument for a
hydraulically explicit model, and it is the reason `g1_eff` is reported at all.

## Light and VPD together

```{r}
ppfds <- seq(50, 1800, length.out = 30)
grid <- do.call(rbind, lapply(c(0.5, 1, 2, 4), function(v) {
  d <- do.call(rbind, lapply(ppfds, function(q) {
    solve_at(psi_soil = 1.0, vpd = v, ppfd = q)
  }))
  d$ppfd <- ppfds
  d
}))

vpds <- sort(unique(grid$vpd))
cols <- hcl.colors(length(vpds), "Zissou 1")
plot(NA, xlim = range(grid$ppfd), ylim = range(grid$A),
     xlab = expression(PPFD~(mu*mol~m^-2~s^-1)),
     ylab = expression(A~(mu*mol~m^-2~s^-1)),
     main = "light response at four vapour pressure deficits")
for (i in seq_along(vpds)) {
  d <- grid[grid$vpd == vpds[i], ]
  lines(d$ppfd, d$A, lwd = 2, col = cols[i])
}
legend("bottomright", legend = paste0("VPD ", vpds, " kPa"),
       col = cols, lwd = 2, bty = "n")
```

The curves saturate at different heights because the hydraulic cost of the
transpiration needed to support a given assimilation rises with VPD — again,
derived, not fitted.

## Where to go next

- `vignette` aside, the C++ API is documented with Doxygen; the model is a set
  of self-contained headers that need no R at all, so the same solve is
  available from a C++ program or a Python extension. See the README.
- `COMPARISON.md` sets this against `plantecophys`, `tealeaves` and `bigleaf`.
- The multi-layer soil path takes vectors for `psi_soil`, `soil_depth` and
  `root_carbon_per_leaf_area`; everything above used a single layer for
  brevity.
