---
title: "The models, and the maths behind them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The models, and the maths behind them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 120, fig.align = "center")
```

```{r setup, message = FALSE}
library(phylloptim)
```

`vignette("phylloptim")` shows the model running; `vignette("fitting")` fits it to
data. This one is the reference for *what is being solved* — the objective, the
cost curves that distinguish one published model from another, the first-order
conditions, and the two gates that change both sides of the ledger.

Every number quoted in prose here is produced by the chunk beside it. That is
deliberate: a formula written next to code it no longer matches is the failure
mode this document is most exposed to.

## The problem

Every model in this package maximises a **profit** — a carbon gain minus a cost
of the water used to get it — over a single water potential:

$$\psi^{*}=\arg\max_{\psi\in[\psi_{\rm soil},\,\psi_{\rm crit}]} J(\psi),
\qquad J(\psi)=B(\psi)-C(\psi)$$

What varies between models is `C`. What varies between *supply topologies* is
which ψ is the decision variable — and that is a bigger difference than it looks.

### Two decision variables, not one

- **Non-root-based.** Optimise the stem potential `ψ_stem` with the upstream
  potential held at `ψ_soil`. The soil-to-collar path is ignored entirely. This is
  what `optimise_psi_stem_*` do, and what their refusal message means by
  "non-root-based".
- **Root-based.** Optimise the **collar** potential with the root path in series,
  then map it to a stem potential. This is `find_root_collar_psi()`.

These are **different models, not one restricted to the other**, and neither is a
check on the other. Measured over 30 driver rows they disagree on 20, and the gap
does not close as the root resistance goes to zero — because as that resistance
vanishes the collar loses its freedom, its feasible interval collapses, and the
root-based solve correctly reports that feasibility *determined* the answer rather
than optimisation.

### The benefit, and the root-find inside it

`B(ψ) = A(c_i)` is colimited assimilation, and `c_i` is not free: it is fixed by
supply meeting demand, which is a root-find at every candidate ψ. Two consequences
run through everything below.

First, `dA/dψ` needs the implicit function theorem rather than a direct
derivative — the chain runs `A'(c_i) · dc_i/dψ`.

Second, there is a regime where that theorem does not hold. When the leaf is hot
enough that assimilation is negative across the whole `[Γ*, c_a]` bracket there is
no supply==demand root, `c_i` is pinned at the compensation point, and **the
residual is no longer zero**. Derivatives on that branch drop the assimilation
term rather than approximating it, and report themselves as not feasible. A
gradient that ignored the flag would get a plausible number there.

## The cost curves

Write the conductivity fraction and its integral as

$$f(\psi)=2^{-(\psi/P_{50})^{c}},\qquad
F(\psi)=\int_0^{\psi}\!f,\qquad
k(\psi)=k_{\max}f(\psi),\qquad
E(\psi;\psi_s)=k_{\max}\big[F(\psi)-F(\psi_s)\big]$$

so `f` is the fraction of maximum conductivity remaining, `k` the conductance and
`E` the transpiration. Appendix A derives this parameterisation and why it is
`(P50, c)` rather than `(b, c)`.

| model | cost `C(ψ)` | marginal cost `dC/dψ` | units of `C` |
|---|---|---|---|
| **TF24** | $\gamma\,(1-f)^{\beta}$ | $\gamma\beta(1-f)^{\beta-1}\lvert f'\rvert$ | carbon |
| **Cowan & Farquhar (1977)** | $\lambda\,E$ | $\lambda\,k_{\max}f$ | carbon |
| **Sperry (2017) ProfitMax** | $\dfrac{k(\psi_s)-k(\psi)}{k(\psi_s)-k_{\rm crit}}$ | $\dfrac{k_{\max}\lvert f'\rvert}{k(\psi_s)-k_{\rm crit}}$ | dimensionless |

The marginal costs are where the models actually differ, and the difference is
one of **shape**:

$$\lvert f'\rvert=\frac{c}{P_{50}}\Big(\frac{\psi}{P_{50}}\Big)^{c-1}f\ln 2$$

which for `c > 1` vanishes as `ψ → 0` and again as `ψ → ∞` — it is peaked. So
Cowan-Farquhar prices water at a marginal rate `∝ f`, falling monotonically as the
leaf dries, while the conductance-based costs price it at `∝ |f'|`, which *rises*
from zero before it falls.

```{r}
l <- leaf_model()
psi <- seq(0.02, l$psi_crit, length.out = 600)
f <- vapply(psi, l$proportion_of_conductivity, numeric(1))
h <- 1e-4
fp <- abs(vapply(psi, function(p) {
  (l$proportion_of_conductivity(p + h) - l$proportion_of_conductivity(p - h)) / (2 * h)
}, numeric(1)))
cat(sprintf("f  is monotone decreasing over the bracket: %s\n",
            all(diff(f) < 0)))
cat(sprintf("|f'| peaks at %.2f MPa; %.1f%% of peak at the wet end, %.0f%% at psi_crit\n",
            psi[which.max(fp)], 100 * fp[1] / max(fp),
            100 * fp[length(fp)] / max(fp)))
```

⚠️ **Read that dry-end figure before generalising.** `|f'|` is asymptotically zero
at both ends, but the bracket stops at `psi_crit`, and there it is still a sizeable
fraction of its peak. So *within the operating range* the shape is "rises from
nothing, then falls somewhat" — the vanishing that matters in practice is at the
**wet** end, which is where pinned optima actually accumulate.

That is still enough to matter: a marginal cost that starts at zero against a
marginal benefit that also saturates is the shape whose first-order condition can
cross zero twice, which is why the conductance forms are the ones needing a search
able to see past a local maximum, and the monotone Cowan-Farquhar cost is not.

## λ is the common currency

Divide any model's marginal cost by the marginal water use and you get the same
kind of quantity: carbon given up per unit of water. That is the **marginal cost
of water**,

$$\lambda(\psi)=\frac{dC/d\psi}{dE/d\psi},$$

and `marginal_cost_water()` reports it for whichever curve is in force. It is the
axis on which otherwise incomparable models can be compared, and the whole content
of the family resemblance between them.

For Cowan-Farquhar it is *by construction* the prescribed constant — that model's
first-order condition is exactly the classical `dA/dE = λ`. For ProfitMax it is
derived, `λ* = |A|max / [k(ψ_s) − k_crit]`, and it is a function of the drivers:
as the soil dries `k(ψ_s)` falls, the denominator shrinks, and water automatically
gets dearer.

**That contrast is the difference between the two models.** A prescribed constant
and a driver-dependent price agree at one operating point and diverge across any
drydown:

```{r}
implied <- vapply(c(0.5, 1, 2, 3), function(ps) {
  x <- leaf_model()
  set_drivers(x, psi_soil = ps, PPFD = 1500)
  x$optimise_psi_stem_TF()
  x$lambda
}, numeric(1))
cat(sprintf(paste0("implied lambda over psi_soil 0.5-3 MPa: ",
                   "%.3g to %.3g umol C / kg, a %.1fx range\n"),
            min(implied), max(implied), max(implied) / min(implied)))
```

So a Cowan-Farquhar λ has to be chosen on the leaf's own scale. A value far below
that band makes water free and pins the optimum wide open; far above it and the
stomata pin shut. Either way the model is not being exercised.

And because both optima satisfy `dA/dE = λ`, handing Cowan-Farquhar the λ that
TF24 *implies* at its own optimum must land it on that same optimum — a check that
needs no derivative approximation anywhere in it:

```{r}
ref <- leaf_model()
set_drivers(ref, psi_soil = 1.0, PPFD = 1500)
ref$optimise_psi_stem_TF()

cf <- leaf_model()
set_drivers(cf, psi_soil = 1.0, PPFD = 1500)
cf$lambda_ <- ref$lambda
cf$optimise_psi_stem_CowanFarquhar()

cat(sprintf(paste0("TF24 optimum %.6f MPa; Cowan-Farquhar at its implied ",
                   "lambda %.6f MPa (gap %.2e)\n"),
            ref$opt_psi_stem_, cf$opt_psi_stem_,
            abs(cf$opt_psi_stem_ - ref$opt_psi_stem_)))
```

## The two gates

Both default off, and both change more than their name suggests.

**Energy balance** (`use_energy_balance_`) makes leaf temperature a function of
transpiration rather than a driver. With it off, a change to any temperature
response is inert at the reference temperature by construction. With it on, `Tleaf`
feeds `vcmax`, `jmax`, `R_d` and the leaf-to-air deficit, and every derivative
picks up a `dT/dE` term.

**Thermal cost** (`use_thermal_cost_`) adds a sigmoid penalty in leaf temperature
— and *also* scales `Jmax` by `(1 − TC)`. So it moves the **benefit** as well as
the cost, which makes it the only gate that changes both sides. It is also inert
without the energy balance, because with a prescribed leaf temperature the penalty
is an additive constant in ψ: it changes the value of the objective and not its
argmax.

## When the leaf cannot transpire

Where the soil is drier than the stem's critical potential there is no flow, and
the operating point is the no-flow one. The answer is to **evaluate the objective
there** rather than to zero the reported fields, so that every output describes one
consistent state. For Cowan-Farquhar this is exactly predictable: `E` is zero, so
the cost is zero, so profit is `−R_d` whatever λ is.

```{r}
d <- leaf_model()                       # psi_soil drier than psi_crit
set_drivers(d, psi_soil = 6.0, PPFD = 1500)
d$lambda_ <- 1.5e5
d$optimise_psi_stem_CowanFarquhar()
cat(sprintf("E = %.1e, cost = %.1e, profit = %.6f, -R_d = %.6f\n",
            d$transpiration_, d$hydraulic_cost_, d$profit_, -d$R_d_))
```

---

# Appendix A — the (P50, c) parameterisation

*Placeholder: the material from `RECOMMENDATION-p50-parameterisation.md` lands
here — the substitution of `b`, why not `(P50, b)`, why `c` is not a constant, the
fitted-species table with its Choat-2003 caveat, the quantile converter, and the
sign convention.*

# Appendix B — gradients

*Placeholder: the implicit-function-theorem composite, `dψ*/dθ = −M/H`, the
envelope theorem and its stationarity condition, and the prescribed-versus-optimum
distinction.*

# Appendix C — numerical practice

*Placeholder: why the argmax is the badly-conditioned quantity and the maximum the
well-conditioned one, the sqrt amplification, the finite-difference step and its
`tol/h` noise floor, and why a scan argmax cannot feed a trait gradient.*
