---
title: "The models, and the maths behind them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The models, and the maths behind them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 120, fig.align = "center")
```

```{r setup, message = FALSE}
library(phylloptim)
```

`vignette("phylloptim")` shows the model running; `vignette("fitting")` fits it to
data. This one is the reference for *what is being solved* — the objective, the
cost curves that distinguish one published model from another, the first-order
conditions, and the two gates that change both sides of the ledger.

Every number quoted in prose here is produced by the chunk beside it. That is
deliberate: a formula written next to code it no longer matches is the failure
mode this document is most exposed to.

## A leaf choosing where to operate

A leaf with open stomata takes up CO₂ and loses water. Opening further buys more
carbon, but the water has to be pulled through the plant at greater tension, and
tension is not free. Somewhere between shut and wide open there is a best place to
sit, and that place is what these models compute.

Put the two effects on the same axis. Take the stem water potential as the dial —
more tension to the right, wider open — and plot what the leaf gains and what it
gives up. That is panel (a) below; the other three panels are the rest of this
document in one frame.

```{r fig-overview, fig.width = 11, fig.height = 6.6, out.width = "100%", dev = "svg"}
l0 <- leaf_model()
set_drivers(l0, psi_soil = 1.0, PPFD = 1500)
plot_model_overview(l0)

l0$optimise_psi_stem_TF()
cat(sprintf("best operating point: %.3f MPa, profit %.3f\n",
            l0$opt_psi_stem_, l0$profit_))
```

Read it in four moves.

- **(a) the trade-off.** The benefit rises and then flattens — pulling harder recruits progressively less conductance, so the carbon return diminishes — while the cost keeps climbing. Their difference therefore has an interior maximum, and the dotted line is where the solver puts it.
- **(b) the root-find inside the benefit.** `c_i` is not free: it sits where diffusive supply meets biochemical demand, and that intersection is recomputed at every candidate ψ.
- **(c) the wiring.** One dial feeds two branches, which meet again at the objective. The two gates are dashed because both default off, and the arrows crossing the gutter are the interactions worth knowing: the energy balance is a *cycle*, and the thermal cost reaches past it into the benefit column.
- **(d) the physical path** all of that describes, with the modelling choices separated from the physics.

**Everything else in this document is a variation on that picture.** The models
differ in the red curve; the supply topologies differ in what the dial is
physically connected to; the two gates change both curves at once.

The figure is `plot_model_overview()`, so it can be regenerated to a vector file
for use outside this vignette:

```{r eval = FALSE}
pdf("overview.pdf", width = 11, height = 6.6)
plot_model_overview()
dev.off()
```

## The problem, stated

$$\psi^{*}=\arg\max_{\psi\in[\psi_{\rm soil},\,\psi_{\rm crit}]} J(\psi),
\qquad J(\psi)=B(\psi)-C(\psi)$$

`B` is net assimilation and `C` the cost of the water used to get it. What varies
between models is `C`; what varies between supply topologies is which ψ the
argmax is taken over.

### The two supply paths

- **Non-root-based.** Optimise the stem potential `ψ_stem` with the upstream
  potential held at `ψ_soil`. The soil-to-collar path is ignored entirely. This is
  what `optimise_psi_stem_*` do, and what their refusal message means by
  "non-root-based".
- **Root-based.** Optimise the **collar** potential with the root path in series,
  then map it to a stem potential. This is `find_root_collar_psi()`.

These are **different models, not one restricted to the other**, and neither is a
check on the other. Measured over 30 driver rows they disagree on 20, and the gap
does not close as the root resistance goes to zero — because as that resistance
vanishes the collar loses its freedom, its feasible interval collapses, and the
root-based solve correctly reports that feasibility *determined* the answer rather
than optimisation.

### One problem, several coordinates

Reading across this literature you will have noticed that different papers
optimise with respect to different things. @cowan1977 solve for stomatal
conductance. @sperry2017 solves for water potential. Others work in terms of
transpiration, or of internal CO₂. It is easy to read that as a substantive
disagreement about what the plant controls.

For the class of models here, it is not. **These variables are interchangeable,
and any of the models could be solved in any of them.** The choice is one of
convenience, not of biology.

The reason is that, given the drivers, the quantities are not independent — they
are linked by a chain of **monotone** maps:

$$\psi \;\xrightarrow{\;\text{supply}\;}\; E \;\xrightarrow{\;\text{diffusion}\;}\; g_c
\;\xrightarrow{\;\text{supply}=\text{demand}\;}\; c_i \;\xrightarrow{\;A(\cdot)\;}\; A$$

Fix any one and the rest follow. Because each map is monotone, it carries the
maximum to the maximum: an optimum located in one coordinate *is* the optimum in
all of them. So "optimise over stem potential", "optimise over stomatal
conductance" and "optimise over `c_i`" name one problem written three ways, and a
model's operating point does not depend on which was chosen.

This package solves in water potential, because that is the coordinate in which
the hydraulic supply constraint is simplest to state. That is an implementation
choice and nothing more — what actually distinguishes the models is the cost
function, which is the next section.

⚠️ **Monotone, not linear** — and that limits how far the equivalence goes. A
*derivative* does depend on the coordinate: `dA/dE` and `dA/dψ` are different
numbers, related by `dE/dψ`. So a marginal cost is only meaningful once you say
what it is per unit of, and comparing a λ from one paper with a λ from another
means checking they share a denominator. `marginal_cost_water()` reports per unit
`E`, which is the convention the λ literature uses.

### Solving for the optimum by its first-order condition

There are two ways to find `ψ*`: search `J` for its maximum, or solve

$$\frac{dJ}{d\psi}=0$$

for a stationary point. This package does both, in different places, and the
difference matters more than it looks.

Solving the first-order condition is better where it applies. It locates the
optimum to solver precision rather than to the width of a search bracket, and —
decisively — it gives an argmax that varies *smoothly* with the parameters. A
search that terminates on bracket width returns an argmax that is piecewise
constant at fine scales, so differentiating it with respect to a trait yields
noise: measured on this model, a trait derivative obtained that way came back
smooth, plausible and **sign-inverted**. Anything that feeds a gradient must
therefore come from the first-order condition, not from a search.

⚠️ **But a stationary point is not always the answer.** Where profit is still
climbing as it reaches the end of the feasible interval, `dJ/dψ` never crosses zero
inside it and the optimum is that endpoint — a *constrained* optimum. This is not
a corner case: it is 42 of 240 feasible rows on the reference grid. A solver has to
test the derivative's sign at each end and report the pin explicitly, because a
bracketing root-find handed no sign change cannot distinguish "pinned" from
"failed". The reported operating-point kind is what carries that distinction.

⚠️ **And a stationary point need not be unique.** Where the marginal cost is
peaked (see the cost curves below) the condition can have two roots, and a
bracketing solve finds one of them. Searching the objective over a grid finds the
taller hump but cannot feed a gradient; solving the condition feeds a gradient but
can land on the wrong hump. The two properties are in tension, and which solver a
given entry point uses reflects which property it needs.

### The benefit, and the root-find inside it

`B(ψ) = A(c_i)` is colimited assimilation, and `c_i` is not free: it is fixed by
supply meeting demand, which is a root-find at every candidate ψ. Two consequences
run through everything below.

First, `dA/dψ` needs the implicit function theorem rather than a direct
derivative — the chain runs `A'(c_i) · dc_i/dψ`.

Second, there is a regime where that theorem does not hold. When the leaf is hot
enough that assimilation is negative across the whole `[Γ*, c_a]` bracket there is
no supply==demand root, `c_i` is pinned at the compensation point, and **the
residual is no longer zero**. Derivatives on that branch drop the assimilation
term rather than approximating it, and report themselves as not feasible. A
gradient that ignored the flag would get a plausible number there.

## The cost curves

Write the conductivity fraction and its integral as

$$f(\psi)=2^{-(\psi/P_{50})^{c}},\qquad
F(\psi)=\int_0^{\psi}\!f,\qquad
k(\psi)=k_{\max}f(\psi),\qquad
E(\psi;\psi_s)=k_{\max}\big[F(\psi)-F(\psi_s)\big]$$

so `f` is the fraction of maximum conductivity remaining, `k` the conductance and
`E` the transpiration. Appendix A derives this parameterisation and why it is
`(P50, c)` rather than `(b, c)`.

| model | cost `C(ψ)` | marginal cost `dC/dψ` | units of `C` |
|---|---|---|---|
| **TF24** [@towers2024] | $\gamma\,(1-f)^{\beta}$ | $\gamma\beta(1-f)^{\beta-1}\lvert f'\rvert$ | carbon |
| **Cowan-Farquhar** [@cowan1977] | $\lambda\,E$ | $\lambda\,k_{\max}f$ | carbon |
| **ProfitMax** [@sperry2017] | $\dfrac{k(\psi_s)-k(\psi)}{k(\psi_s)-k_{\rm crit}}$ | $\dfrac{k_{\max}\lvert f'\rvert}{k(\psi_s)-k_{\rm crit}}$ | dimensionless |

The marginal costs are where the models actually differ, and the difference is
one of **shape**:

$$\lvert f'\rvert=\frac{c}{P_{50}}\Big(\frac{\psi}{P_{50}}\Big)^{c-1}f\ln 2$$

which for `c > 1` vanishes as `ψ → 0` and again as `ψ → ∞` — it is peaked. So
Cowan-Farquhar prices water at a marginal rate `∝ f`, falling monotonically as the
leaf dries, while the conductance-based costs price it at `∝ |f'|`, which *rises*
from zero before it falls.

```{r}
l <- leaf_model()
psi <- seq(0.02, l$psi_crit, length.out = 600)
f <- vapply(psi, l$proportion_of_conductivity, numeric(1))
h <- 1e-4
fp <- abs(vapply(psi, function(p) {
  (l$proportion_of_conductivity(p + h) - l$proportion_of_conductivity(p - h)) / (2 * h)
}, numeric(1)))
cat(sprintf("f  is monotone decreasing over the bracket: %s\n",
            all(diff(f) < 0)))
cat(sprintf("|f'| peaks at %.2f MPa; %.1f%% of peak at the wet end, %.0f%% at psi_crit\n",
            psi[which.max(fp)], 100 * fp[1] / max(fp),
            100 * fp[length(fp)] / max(fp)))
```

⚠️ **Read that dry-end figure before generalising.** `|f'|` is asymptotically zero
at both ends, but the bracket stops at `psi_crit`, and there it is still a sizeable
fraction of its peak. So *within the operating range* the shape is "rises from
nothing, then falls somewhat" — the vanishing that matters in practice is at the
**wet** end, which is where pinned optima actually accumulate.

That is still enough to matter: a marginal cost that starts at zero against a
marginal benefit that also saturates is the shape whose first-order condition can
cross zero twice, which is why the conductance forms are the ones needing a search
able to see past a local maximum, and the monotone Cowan-Farquhar cost is not.

## λ is the common currency

Divide any model's marginal cost by the marginal water use and you get the same
kind of quantity: carbon given up per unit of water. That is the **marginal cost
of water**,

$$\lambda(\psi)=\frac{dC/d\psi}{dE/d\psi},$$

and `marginal_cost_water()` reports it for whichever curve is in force. It is the
axis on which otherwise incomparable models can be compared, and the whole content
of the family resemblance between them.

For Cowan-Farquhar it is *by construction* the prescribed constant — that model's
first-order condition is exactly the classical `dA/dE = λ`. For ProfitMax it is
derived, `λ* = |A|max / [k(ψ_s) − k_crit]`, and it is a function of the drivers:
as the soil dries `k(ψ_s)` falls, the denominator shrinks, and water automatically
gets dearer.

**That contrast is the difference between the two models.** A prescribed constant
and a driver-dependent price agree at one operating point and diverge across any
drydown:

```{r}
implied <- vapply(c(0.5, 1, 2, 3), function(ps) {
  x <- leaf_model()
  set_drivers(x, psi_soil = ps, PPFD = 1500)
  x$optimise_psi_stem_TF()
  x$lambda
}, numeric(1))
cat(sprintf(paste0("implied lambda over psi_soil 0.5-3 MPa: ",
                   "%.3g to %.3g umol C / kg, a %.1fx range\n"),
            min(implied), max(implied), max(implied) / min(implied)))
```

So a Cowan-Farquhar λ has to be chosen on the leaf's own scale. A value far below
that band makes water free and pins the optimum wide open; far above it and the
stomata pin shut. Either way the model is not being exercised.

And because both optima satisfy `dA/dE = λ`, handing Cowan-Farquhar the λ that
TF24 *implies* at its own optimum must land it on that same optimum — a check that
needs no derivative approximation anywhere in it:

```{r}
ref <- leaf_model()
set_drivers(ref, psi_soil = 1.0, PPFD = 1500)
ref$optimise_psi_stem_TF()

cf <- leaf_model()
set_drivers(cf, psi_soil = 1.0, PPFD = 1500)
cf$lambda_ <- ref$lambda
cf$optimise_psi_stem_CowanFarquhar()

cat(sprintf(paste0("TF24 optimum %.6f MPa; Cowan-Farquhar at its implied ",
                   "lambda %.6f MPa (gap %.2e)\n"),
            ref$opt_psi_stem_, cf$opt_psi_stem_,
            abs(cf$opt_psi_stem_ - ref$opt_psi_stem_)))
```

### λ is not water use efficiency

The two are easy to conflate: same units, both "carbon per water". They are
different quantities, and the difference is marginal against average.

$$\text{WUE}=\frac{A}{E}\quad\text{(an average)},
\qquad
\lambda=\frac{dA}{dE}\quad\text{(a marginal rate)}$$

WUE is the slope of the chord from the origin to the operating point; λ is the
slope of the tangent there. On a saturating `A(E)` those differ, and the gap is
not a fixed factor:

```{r}
cat(" psi_soil   lambda      A/E (WUE)   lambda/WUE\n")
for (ps in c(0.5, 1, 2, 3)) {
  lw <- leaf_model()
  set_drivers(lw, psi_soil = ps, PPFD = 1500)
  lw$optimise_psi_stem_TF()
  cat(sprintf(" %7.1f  %10.4g  %10.4g  %9.3f\n", ps, lw$lambda_emergent,
              lw$assim_colimited_ / lw$transpiration_,
              lw$lambda_emergent / (lw$assim_colimited_ / lw$transpiration_)))
}
```

So λ runs a third to three-quarters of WUE here, and the ratio **rises as the soil
dries** — you cannot infer one from the other, in either direction.

⚠️ **What is true is narrower, and only at the optimum.** The first-order
condition is `dA/dE = λ`, so at the operating point λ *is* the marginal water use
efficiency. Off the optimum it is not even that. And "WUE" unqualified almost
always means the average `A/E`, or intrinsic WUE `A/g_s` — so a λ from this model
is not comparable with a WUE from the literature without checking which was meant.

## The two gates

Both default off, and both change more than their name suggests.

**Energy balance** (`use_energy_balance_`) makes leaf temperature a function of
transpiration rather than a driver. With it off, a change to any temperature
response is inert at the reference temperature by construction. With it on, `Tleaf`
feeds `vcmax`, `jmax`, `R_d` and the leaf-to-air deficit, and every derivative
picks up a `dT/dE` term.

**Thermal cost** (`use_thermal_cost_`) adds a sigmoid penalty in leaf temperature
— and *also* scales `Jmax` by `(1 − TC)`. So it moves the **benefit** as well as
the cost, which makes it the only gate that changes both sides. It is also inert
without the energy balance, because with a prescribed leaf temperature the penalty
is an additive constant in ψ: it changes the value of the objective and not its
argmax.

## When the leaf cannot transpire

Where the soil is drier than the stem's critical potential there is no flow, and
the operating point is the no-flow one. The answer is to **evaluate the objective
there** rather than to zero the reported fields, so that every output describes one
consistent state. For Cowan-Farquhar this is exactly predictable: `E` is zero, so
the cost is zero, so profit is `−R_d` whatever λ is.

```{r}
d <- leaf_model()                       # psi_soil drier than psi_crit
set_drivers(d, psi_soil = 6.0, PPFD = 1500)
d$lambda_ <- 1.5e5
d$optimise_psi_stem_CowanFarquhar()
cat(sprintf("E = %.1e, cost = %.1e, profit = %.6f, -R_d = %.6f\n",
            d$transpiration_, d$hydraulic_cost_, d$profit_, -d$R_d_))
```

---

## The variants, under one roof

Four axes, chosen independently. A "model" in the literature is a point in this
space, not a separate program.

| axis | options | what it changes |
|---|---|---|
| **cost curve** | TF24 / Cowan-Farquhar / ProfitMax | `C(ψ)`, and its units |
| **supply topology** | single soil potential / multi-layer root network | the decision variable, and the feasible bracket |
| **leaf temperature** | prescribed / energy balance | whether `Tleaf` responds to `E`, and so whether temperature responses are active at all |
| **thermal cost** | off / on | adds a temperature penalty **and** scales `Jmax` — the only axis that moves both sides |

So the published models map on as follows:

- **Cowan & Farquhar** [@cowan1977] — the `λE` cost, prescribed λ. Any topology.
- **Sperry et al.** [@sperry2017] — ProfitMax cost with the normalisation, λ
  derived per operating point. Their own formulation is single-supply.
- **Towers et al.** [@towers2024] ("TF24") — the TF24 carbon cost, solved over the
  collar potential with a multi-layer root network. This is the production path
  here, and the one `plant` consumes.
- **Sicangco et al.** [@sicangco2026] — ProfitMax with the thermal cost on.
- **Jones et al.** [@jones2026] — a multiplicative objective; see the note below.

⚠️ **Jones et al. is not yet in this list as an implemented cost curve.** Its
objective is a **product**, `A · (1 − ψ_leaf/ψ_crit)`, not a difference. Taking
logs turns it into one — `log J = log A + log(1 − ψ/ψ_crit)` — which is the same
additive skeleton with a `log` benefit link, and the argmax is unchanged because
`log` is monotone. Two constraints come with it: `log A` requires `A > 0`, which
fails exactly in the closed-stomata state where `A = −R_d`; and the cost diverges
at `ψ_crit`, so a Jones-type model can never be pinned dry. Both have to be
enforced rather than allowed to become `NaN`.

## The biological components

The optimality argument above is indifferent to what `A` and `E` actually are.
What makes this a plant model rather than a calculus exercise is the five blocks
underneath, given here in the forms the code evaluates.

### 1. Photosynthetic demand

Farquhar-type assimilation [@farquhar1980]. A Rubisco-limited rate and an
electron-transport-limited rate, each a saturating function of `c_i`:

$$A_c=\frac{V_{c\max}\,(c_i-\Gamma^{*})}{c_i+K_m},
\qquad
A_j=\frac{J}{4}\cdot\frac{c_i-\Gamma^{*}}{c_i+2\Gamma^{*}}$$

Electron transport itself is a smooth minimum of light supply and capacity:

$$J=\frac{aQ+J_{\max}-\sqrt{(aQ+J_{\max})^{2}-4\,\theta_J\,aQ\,J_{\max}}}{2\theta_J}$$

and the two rates are colimited by the *same* smooth-minimum form before
respiration is subtracted:

$$A=\frac{A_c+A_j-\sqrt{(A_c+A_j)^{2}-4\,\theta_c A_c A_j}}{2\theta_c}-R_d$$

`θ_J` and `θ_c` are curvature factors — at 1 the expression is an exact minimum,
below 1 it rounds the corner. The Michaelis-Menten composite carries the
competitive inhibition by oxygen:

$$K_m=K_c\Big(1+\frac{O_2}{K_o}\Big)$$

⚠️ **`A` here is NET assimilation.** `R_d` is inside it, which is why the
closed-stomata profit is `−R_d` rather than zero, and why `A` can be negative.

### 2. Temperature responses

`V_cmax`, `J_max`, `Γ*`, `K_c`, `K_o` and `R_d` are all defined at 25 °C and moved
from there. The first two get a **peaked** Arrhenius — rising, then falling as
denaturation takes over:

$$X(T)=X_{25}\exp\!\Big[\frac{E_a\,(T-T_{25})}{T_{25}\,R\,T}\Big]
\cdot\frac{1+\exp\!\big[(\Delta S\,T_{25}-H_d)/(R\,T_{25})\big]}
          {1+\exp\!\big[(\Delta S\,T-H_d)/(R\,T)\big]}$$

with `T` in kelvin. `Γ*`, `K_c` and `K_o` take the first factor alone. Respiration
is different again — a Q₁₀ whose own value declines with temperature:

$$R_d=R_{d,25}\;q^{(T-25)/10},\qquad q=q_0-q_1\frac{T+25}{2}$$

⚠️ **This is why a grid at 25 °C is blind to every one of these by construction**:
each response is *defined* to be 1 at its reference, so a change to any of them is
inert there. Any test of a temperature response needs a second temperature.

⚠️ **And the thermal-cost gate reaches into this block**: when it is on,
`J_max` is multiplied by `(1 − TC)`. That is why it moves the benefit and not only
the cost.

### 3. Diffusive supply

CO₂ in and water out share the stomata, so one conductance sets both. Converting
the mass flux `E` to a molar CO₂ conductance:

$$g_c=\frac{P\,E\,M_{\rm H_2O}}{1.6\;D_{\rm leaf}}$$

where 1.6 is the ratio of diffusivities of water vapour and CO₂, `P` is
atmospheric pressure and `D_leaf` the leaf-to-air vapour pressure deficit. Fixing
`E` fixes `g_c`, which is the first link of the coordinate chain above.

⚠️ **`$stom_cond_CO2_` is a TOTAL leaf-to-air conductance, not a stomatal one.**
There is no separate boundary-layer term: the relation above is the whole path from
leaf interior to bulk air, in one conductance, paired with a leaf-to-bulk-air
deficit. That is internally consistent, and it is the right quantity for the `c_i`
solve below. But a measured `g_s` has a boundary layer removed from it, so the two
are not the same number and should not be compared directly.

### 4. Closing the loop: supply equals demand

`c_i` is not free. Diffusive supply must equal biochemical demand, which is a
scalar root-find at every candidate ψ:

$$g(c_i)\;=\;A(c_i)\;-\;\frac{g_c\,(c_a-c_i)}{P}\;=\;0$$

solved over `[Γ*, c_a]` by a bracketing method. This is the innermost loop of the
whole model — it runs at every objective evaluation, and its tolerance sets what
every reported output *means*.

```{r}
# The residual really is zero at the solved ci -- the statement above, checked.
l2 <- leaf_model()
set_drivers(l2, psi_soil = 1.0, PPFD = 1500)
l2$optimise_psi_stem_TF()
ci <- l2$ci_
gc <- l2$stom_cond_CO2_
resid <- l2$assim_colimited(ci) * 1e-6 - gc * (l2$ca_ - ci) / (l2$atm_kpa_ * 1000)
cat(sprintf("ci = %.4f Pa, supply-demand residual = %.2e mol m-2 s-1\n",
            ci, resid))
```

### 5. Hydraulic supply, and why the problem is non-trivial

Water arrives through a path whose conductivity *falls* as tension rises — the
Weibull vulnerability curve of Appendix A. Transpiration at a given stem potential
is that curve integrated along the path:

$$E(\psi_{\rm stem};\psi_{\rm up})=k_{\max}\int_{\psi_{\rm up}}^{\psi_{\rm stem}}\! f(u)\,du$$

**This integral is what makes the optimisation non-trivial.** Because `f` falls,
`E` *saturates*: pulling harder recruits progressively less conductance, so the
carbon return on further tension diminishes. Without that saturation more tension
would always buy more carbon and there would be no interior optimum to find.

### 6. Optionally, energy balance and thermal damage

A transpiring leaf is cooler than the air. Linearised, the leaf temperature is

$$T_{\rm leaf}=T_{\rm air}+\frac{(R_n-\lambda_v E)\,r_a}{\rho c_p}$$

clamped to a physical range — so `dT/dE = −λ_v r_a/(\rho c_p)` inside the range and
exactly zero where the clamp binds, which is a real kink every derivative has to
respect. Warming the leaf also raises its own vapour pressure, so the deficit that
drives diffusion is not the atmospheric one:

$$D_{\rm leaf}=\max\!\big(D_{\rm air}+e_s(T_{\rm leaf})-e_s(T_{\rm air}),\;D_{\min}\big)$$

That closes a **feedback**: `E` sets `T_leaf`, which sets `D_leaf`, which sets
`g_c`, which sets `E`. It is why the energy-balance path is solved rather than
evaluated, and why every derivative on it carries a `dT/dE` term.

Finally, the thermal cost — a sigmoid in leaf temperature between a critical and a
half-lethal value [@sicangco2026]:

$$TC(T)=\frac{1}{1+\exp\!\big[-r\,(T-T_{50})\big]},\qquad r=\frac{2}{T_{50}-T_{\rm crit}}$$

With this on, a leaf has a reason to transpire that is not photosynthetic at all:
evaporative cooling becomes a benefit in its own right.

### What the models disagree about

Blocks 1–5 are shared. Every published model in the table above accepts them and
differs only in block 6's cost: a carbon penalty rising with the fraction of
conductivity lost [@towers2024], the water itself priced at a constant marginal
rate [@cowan1977], or the normalised loss of remaining conductance [@sperry2017].
They are **not refinements of one another** — they disagree about what is being
spent.

### Further reading

@sabot2022 compare several of these models in one notation, and are the closest
published analogue to the framing used here. @medlyn2011 give the empirical
conductance model this package can report an equivalent slope for. Full citations
are in the reference list at the end.

# Appendix A — the (P50, c) parameterisation

The Weibull vulnerability curve is conventionally written `exp(-(ψ/b)^c)`. Substituting
`b = P50 / (ln 2)^(1/c)` gives the algebraically identical

$$f(\psi)=2^{-(\psi/P_{50})^{c}}$$

in which **scale and shape are cleanly separated**: `P50` is the potential at 50%
loss, in MPa, and `c` is the steepness. `b` remains available as a derived
read-only quantity, but it is not a parameter anyone should set.

⚠️ **Sign convention.** Published `P50` values are negative (−3.95 MPa). This
package uses positive magnitudes throughout, so `P50` here means `|P50|`.

## Why not `(P50, b)`

`b` *is* a scale parameter — it is the 63% loss point, in the same units and the
same role as `P50`. Specifying both means specifying two points on one curve and
recovering the shape from their ratio:

$$c=\frac{\ln(\ln 2)}{\ln(P_{50}/b)}$$

which diverges as `P50 → b` and is undefined at `P50 = b`, so small errors in
either input make the shape explode. It is also a badly correlated estimation
pair — which is why `fitplc` [@duursma2017] fits `Px` and the *slope* at `Px`
rather than `b` and `c`. `c` is the only genuinely independent second axis.

## Why `c` is not a constant

It carries real biology: `c ≈ 1` is exponential, conductivity bleeding away from
`ψ = 0`; `c > 2` is sigmoid, a safe plateau then rapid failure. Conifers are much
steeper than angiosperms, and leaves and roots are flatter than stems in the same
plant.

Fitted directly in `(P50, c)` from raw PLC curves for eight species:

| species | source / method | \|P50\| (MPa) | c | S50 (%/MPa) |
|---|---|---|---|---|
| *Eucalyptus sideroxylon* | Blackman 2019, bench | 3.95 | 1.99 | 17 |
| *Eucalyptus tereticornis* | Blackman 2019, bench | 4.31 | 2.59 | 21 |
| *Casuarina cunninghamiana* | Blackman 2019, bench | 2.91 | 1.81 | 22 |
| *Pinus radiata* | Blackman 2019, cavitron | 4.09 | 13.0 | 110 |
| *Gossia bidwillii* | Choat 2003 | 5.25 | 0.52 | 3 |
| *Brachychiton australis* | Choat 2003 | 3.49 | 1.21 | 12 |
| *Alphitonia excelsa* | Choat 2003 | 5.69 | 1.70 | 10 |
| *Cochlospermum gillivraei* | Choat 2003 | 1.42 | 5.15 | 126 |

⚠️ **The four Choat 2003 curves are not trustworthy for `c`.** They are 13–14
points each and already sit at 25–30% PLC by 0.8 MPa, which is a native-embolism
or open-vessel artefact rather than a trait. The defensible comparison is the
three Blackman angiosperms measured the same way: **`c` = 1.8–2.6 while `|P50|`
spans 2.9–4.3 MPa** — shape roughly conserved, scale free. That is the evidence
for treating `c` as a parameter with a stable default rather than a constant.

Default `c = 2.0` for angiosperm stems. Conifers need substantially higher; the
13.0 fitted above is inflated by the cavitron technique, and ~4–5 is the safer
literature-style default. Guard `c > 0`, and warn below ~1 — that means no safe
plateau at all, and in practice usually flags a measurement artefact.

## Converting from published numbers

Curves arrive as almost anything except `(P50, c)`. Everything inverts in closed
form. For a remaining conductivity fraction `f` (so `PLC = 1 − f`):

$$P_f = P_{50}\big(\log_2(1/f)\big)^{1/c}
\qquad\Longleftrightarrow\qquad
P_{50} = P_f\big(\log_2(1/f)\big)^{-1/c}$$

From **any two** quantiles, both parameters follow:

$$c=\frac{\ln(L_2/L_1)}{\ln(P_{f_2}/P_{f_1})},\qquad L_i=\log_2(1/f_i)$$

And from `P50` plus the slope at `P50` in % MPa⁻¹, the other commonly published
form:

$$S_{50}=\frac{50\,c\ln 2}{|P_{50}|}=\frac{34.657\,c}{|P_{50}|}$$

Do **not** treat `S50` as the internal parameter — it has `P50` built into it by
construction.

```{r}
L <- function(f) log2(1 / f)                    # f = fraction REMAINING
p_at <- function(P50, c, f) P50 * L(f)^(1 / c)  # any quantile from (P50, c)
c_from_two <- function(P1, f1, P2, f2) log(L(f2) / L(f1)) / log(P2 / P1)
s50_from <- function(P50, c) 50 * c * log(2) / P50

# round-trip: the package defaults, recovered from a (P50, P88) pair
tr <- leaf_traits()
l  <- leaf_model()
P88 <- p_at(tr$stem_P50, tr$stem_c, 0.12)
c_hat <- c_from_two(tr$stem_P50, 0.50, P88, 0.12)
P50_hat <- P88 * L(0.12)^(-1 / c_hat)
cat(sprintf("defaults: P50 = %.6f MPa, c = %.6f, so P88 = %.6f, P95 = %.6f\n",
            tr$stem_P50, tr$stem_c, P88, p_at(tr$stem_P50, tr$stem_c, 0.05)))
cat(sprintf("recovered from (P50, P88): c = %.6f, P50 = %.6f  (errors %.1e, %.1e)\n",
            c_hat, P50_hat, abs(c_hat - tr$stem_c), abs(P50_hat - tr$stem_P50)))
cat(sprintf("S50 at the defaults: %.1f %%/MPa\n", s50_from(tr$stem_P50, tr$stem_c)))

# the derived pair, at FULL precision -- b and psi_crit are read off the model,
# the right-hand sides are this appendix's own formulae
cat(sprintf("stem_b   = %.17g  vs  P50/(ln2)^(1/c) = %.17g\n",
            l$stem_b, tr$stem_P50 / log(2)^(1 / tr$stem_c)))
cat(sprintf("psi_crit = %.17g  vs  P95            = %.17g\n",
            l$psi_crit, p_at(tr$stem_P50, tr$stem_c, 0.05)))
cat(sprintf("f(psi_crit) = %.17g   (the critical fraction, exactly)\n",
            l$proportion_of_conductivity(l$psi_crit)))
```

Those are equalities, not agreements: `stem_b` and `psi_crit` are computed from
`(P50, c)` by the expressions printed beside them, so there is no gap to quote.

⚠️ **It was not always so, and the difference is worth recording.** The defaults
used to be round values of `stem_b` and `psi_crit` rather than of `P50`, with
`psi_crit` set to `P95` *rounded to seven figures* — so `f(psi_crit)` was
`0.049999968739194864` rather than `0.05`, and the header claiming it was "0.0500
exactly" was true only to four decimal places. Adopting a round `P50 = 3.4` with
both derived moved `stem_b` by 3.1e-08 and `psi_crit` by 4.7e-08. That propagated
to a worst 4.4e-05 on assimilation across the 576-point reference grid — a
thousandfold amplification of a rounding-level input change, and still below the
1e-4 at which this project calls a difference real.

## `psi_crit` is a quantile, not a parameter

`psi_crit` is `P_f` at `f = 0.05`, i.e. `P95` — and **Sperry's reference
conductivity `k_crit = f·kmax` uses the same `f`**. So `psi_crit`, `P95` and
`k_crit` are one number wearing three names. Derived from a single named constant
they cannot disagree; held as a free trait plus a hard-coded `0.05`, they needed a
consistency check to police them. `f = 0.05` is a convention, not a knob — it is
0.05 everywhere it appears across this family.

Two things follow that are easy to miss:

- The **domain check disappears**, because it becomes vacuous. It enforced
  `psi_crit ≤ P99`, and a derived `psi_crit = P95` satisfies that for every `c > 0`
  since `ln 20 < ln 100`. There is nothing left to validate.
- A **shut-down leaf's hydraulic cost stops depending on the curve's scale.** It
  sits at `psi_crit`, where the remaining conductivity is exactly `f` whatever
  `P50` is — so the cost is the same number for every `P50`, and
  `dprofit/dP50` at a shut-down point is exactly zero rather than merely small.

## Three consequences for gradients

All three are asserted in the suite rather than described here, and the numbers
below are what those assertions compare against.

1. **The homogeneity shortcut survives.** At fixed `c`, `b` is a constant
   multiple of `P50`, so the cumulative-vulnerability integral stays homogeneous of
   degree 1 and `d/dP50 = (ln 2)^{-1/c} · d/db` — a factor of 1.1465 at the
   defaults. The existing fast path is reused with that scalar rather than
   rewritten. `psi_crit` scales by the same factor, which is why moving the trait
   can never walk the solve off the end of its own curve.
2. **The `c` gradient splits.** Once `P50` is the scale parameter, `b` depends on
   `c` too:
   $$\frac{d}{dc}\bigg|_{P_{50}}=\frac{d}{dc}\bigg|_{b}+\frac{db}{dc}\frac{d}{db},
   \qquad \frac{db}{dc}=\frac{b\ln(\ln 2)}{c^{2}}$$
   The second term is nearly free — it reuses the cheap `b`-derivative — but
   omitting it silently converts a total derivative into a partial, which looks
   like a plausible-but-wrong gradient rather than an error.
3. ⚠️ **A derived `psi_crit` adds a term.** Moving `P50` moves the threshold, so
   `d/dP50` picks up `dpsi_crit/dP50 · d/dpsi_crit`, with `dpsi_crit/dP50 =
   (log₂(1/f))^{1/c}` — 1.7266 at the defaults, a constant at fixed `c`. That term
   is **exactly zero at an interior optimum and order 1 at a pinned one**, so the
   clean identity in (1) holds *only where the optimum is interior*.

   The size of it is the reason to care. At a dry-pinned point the two halves are
   `1.1465 × 4.6991` from the curve and `1.7266 × 1.2605` through the bound, so
   dropping the second gives 5.39 where the answer is 7.56 — a 29% error with
   nothing about it that looks wrong.

The root curve obeys the identical homogeneity identity and takes the same
treatment — `(root_P50, root_c)` — because two curves on different conventions is
the shape of the mistake that once carried a root parameter into a stem cost.

# Appendix B — gradients

*Placeholder: the implicit-function-theorem composite, `dψ*/dθ = −M/H`, the
envelope theorem and its stationarity condition, and the prescribed-versus-optimum
distinction.*

# Appendix C — numerical practice

*Placeholder: why the argmax is the badly-conditioned quantity and the maximum the
well-conditioned one, the sqrt amplification, the finite-difference step and its
`tol/h` noise floor, and why a scan argmax cannot feed a trait gradient.*
# References
