---
title: "The models, and the maths behind them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The models, and the maths behind them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 120, fig.align = "center")
```

```{r setup, message = FALSE}
library(phylloptim)
```

`vignette("phylloptim")` shows the model running; `vignette("fitting")` fits it to
data. This one is the reference for *what is being solved* — the objective, the
cost curves that distinguish one published model from another, the first-order
conditions, and the two gates that change both sides of the ledger.

Every number quoted in prose here is produced by the chunk beside it. That is
deliberate: a formula written next to code it no longer matches is the failure
mode this document is most exposed to.

## The problem

Every model in this package maximises a **profit** — a carbon gain minus a cost
of the water used to get it — over a single water potential:

$$\psi^{*}=\arg\max_{\psi\in[\psi_{\rm soil},\,\psi_{\rm crit}]} J(\psi),
\qquad J(\psi)=B(\psi)-C(\psi)$$

What varies between models is `C`. What varies between *supply topologies* is
which ψ is the decision variable — and that is a bigger difference than it looks.

### Two decision variables, not one

- **Non-root-based.** Optimise the stem potential `ψ_stem` with the upstream
  potential held at `ψ_soil`. The soil-to-collar path is ignored entirely. This is
  what `optimise_psi_stem_*` do, and what their refusal message means by
  "non-root-based".
- **Root-based.** Optimise the **collar** potential with the root path in series,
  then map it to a stem potential. This is `find_root_collar_psi()`.

These are **different models, not one restricted to the other**, and neither is a
check on the other. Measured over 30 driver rows they disagree on 20, and the gap
does not close as the root resistance goes to zero — because as that resistance
vanishes the collar loses its freedom, its feasible interval collapses, and the
root-based solve correctly reports that feasibility *determined* the answer rather
than optimisation.

### One problem, four coordinates

`ψ`, `E`, `g_c`, `c_i` and `A` are not five independent quantities. Given the
drivers they are linked by a chain of **monotone** maps:

$$\psi \;\xrightarrow{\;\text{supply}\;}\; E \;\xrightarrow{\;\text{diffusion}\;}\; g_c
\;\xrightarrow{\;\text{supply}=\text{demand}\;}\; c_i \;\xrightarrow{\;A(\cdot)\;}\; A$$

Fix any one and the rest follow. So "optimise over stem potential", "optimise over
stomatal conductance" and "optimise over `c_i`" are **the same problem written in
different coordinates**, and a model's argmax does not depend on which is chosen.

That is what makes the literature comparable at all. @cowan1977 write the
problem over conductance; @sperry2017 writes it over potential; this package solves it
over potential because that is where the supply constraint is simplest. None of
that is a modelling difference — only the cost function is.

⚠️ **Monotone, not linear.** A derivative *does* depend on the coordinate: `dA/dE`
and `dA/dψ` are different numbers, related by `dE/dψ`. So a marginal cost is only
meaningful once you say what it is per unit of. `marginal_cost_water()` is per unit
`E`, which is the convention the λ literature uses.

### Solving for the optimum by its first-order condition

There are two ways to find `ψ*`: search `J` for its maximum, or solve

$$\frac{dJ}{d\psi}=0$$

for a stationary point. This package does both, in different places, and the
difference matters more than it looks.

Solving the first-order condition is better where it applies. It locates the
optimum to solver precision rather than to the width of a search bracket, and —
decisively — it gives an argmax that varies *smoothly* with the parameters. A
search that terminates on bracket width returns an argmax that is piecewise
constant at fine scales, so differentiating it with respect to a trait yields
noise: measured on this model, a trait derivative obtained that way came back
smooth, plausible and **sign-inverted**. Anything that feeds a gradient must
therefore come from the first-order condition, not from a search.

⚠️ **But a stationary point is not always the answer.** Where profit is still
climbing as it reaches the end of the feasible interval, `dJ/dψ` never crosses zero
inside it and the optimum is that endpoint — a *constrained* optimum. This is not
a corner case: it is 42 of 240 feasible rows on the reference grid. A solver has to
test the derivative's sign at each end and report the pin explicitly, because a
bracketing root-find handed no sign change cannot distinguish "pinned" from
"failed". The reported operating-point kind is what carries that distinction.

⚠️ **And a stationary point need not be unique.** Where the marginal cost is
peaked (see the cost curves below) the condition can have two roots, and a
bracketing solve finds one of them. Searching the objective over a grid finds the
taller hump but cannot feed a gradient; solving the condition feeds a gradient but
can land on the wrong hump. The two properties are in tension, and which solver a
given entry point uses reflects which property it needs.

### The benefit, and the root-find inside it

`B(ψ) = A(c_i)` is colimited assimilation, and `c_i` is not free: it is fixed by
supply meeting demand, which is a root-find at every candidate ψ. Two consequences
run through everything below.

First, `dA/dψ` needs the implicit function theorem rather than a direct
derivative — the chain runs `A'(c_i) · dc_i/dψ`.

Second, there is a regime where that theorem does not hold. When the leaf is hot
enough that assimilation is negative across the whole `[Γ*, c_a]` bracket there is
no supply==demand root, `c_i` is pinned at the compensation point, and **the
residual is no longer zero**. Derivatives on that branch drop the assimilation
term rather than approximating it, and report themselves as not feasible. A
gradient that ignored the flag would get a plausible number there.

## The cost curves

Write the conductivity fraction and its integral as

$$f(\psi)=2^{-(\psi/P_{50})^{c}},\qquad
F(\psi)=\int_0^{\psi}\!f,\qquad
k(\psi)=k_{\max}f(\psi),\qquad
E(\psi;\psi_s)=k_{\max}\big[F(\psi)-F(\psi_s)\big]$$

so `f` is the fraction of maximum conductivity remaining, `k` the conductance and
`E` the transpiration. Appendix A derives this parameterisation and why it is
`(P50, c)` rather than `(b, c)`.

| model | cost `C(ψ)` | marginal cost `dC/dψ` | units of `C` |
|---|---|---|---|
| **TF24** [@towers2024] | $\gamma\,(1-f)^{\beta}$ | $\gamma\beta(1-f)^{\beta-1}\lvert f'\rvert$ | carbon |
| **Cowan-Farquhar** [@cowan1977] | $\lambda\,E$ | $\lambda\,k_{\max}f$ | carbon |
| **ProfitMax** [@sperry2017] | $\dfrac{k(\psi_s)-k(\psi)}{k(\psi_s)-k_{\rm crit}}$ | $\dfrac{k_{\max}\lvert f'\rvert}{k(\psi_s)-k_{\rm crit}}$ | dimensionless |

The marginal costs are where the models actually differ, and the difference is
one of **shape**:

$$\lvert f'\rvert=\frac{c}{P_{50}}\Big(\frac{\psi}{P_{50}}\Big)^{c-1}f\ln 2$$

which for `c > 1` vanishes as `ψ → 0` and again as `ψ → ∞` — it is peaked. So
Cowan-Farquhar prices water at a marginal rate `∝ f`, falling monotonically as the
leaf dries, while the conductance-based costs price it at `∝ |f'|`, which *rises*
from zero before it falls.

```{r}
l <- leaf_model()
psi <- seq(0.02, l$psi_crit, length.out = 600)
f <- vapply(psi, l$proportion_of_conductivity, numeric(1))
h <- 1e-4
fp <- abs(vapply(psi, function(p) {
  (l$proportion_of_conductivity(p + h) - l$proportion_of_conductivity(p - h)) / (2 * h)
}, numeric(1)))
cat(sprintf("f  is monotone decreasing over the bracket: %s\n",
            all(diff(f) < 0)))
cat(sprintf("|f'| peaks at %.2f MPa; %.1f%% of peak at the wet end, %.0f%% at psi_crit\n",
            psi[which.max(fp)], 100 * fp[1] / max(fp),
            100 * fp[length(fp)] / max(fp)))
```

⚠️ **Read that dry-end figure before generalising.** `|f'|` is asymptotically zero
at both ends, but the bracket stops at `psi_crit`, and there it is still a sizeable
fraction of its peak. So *within the operating range* the shape is "rises from
nothing, then falls somewhat" — the vanishing that matters in practice is at the
**wet** end, which is where pinned optima actually accumulate.

That is still enough to matter: a marginal cost that starts at zero against a
marginal benefit that also saturates is the shape whose first-order condition can
cross zero twice, which is why the conductance forms are the ones needing a search
able to see past a local maximum, and the monotone Cowan-Farquhar cost is not.

## λ is the common currency

Divide any model's marginal cost by the marginal water use and you get the same
kind of quantity: carbon given up per unit of water. That is the **marginal cost
of water**,

$$\lambda(\psi)=\frac{dC/d\psi}{dE/d\psi},$$

and `marginal_cost_water()` reports it for whichever curve is in force. It is the
axis on which otherwise incomparable models can be compared, and the whole content
of the family resemblance between them.

For Cowan-Farquhar it is *by construction* the prescribed constant — that model's
first-order condition is exactly the classical `dA/dE = λ`. For ProfitMax it is
derived, `λ* = |A|max / [k(ψ_s) − k_crit]`, and it is a function of the drivers:
as the soil dries `k(ψ_s)` falls, the denominator shrinks, and water automatically
gets dearer.

**That contrast is the difference between the two models.** A prescribed constant
and a driver-dependent price agree at one operating point and diverge across any
drydown:

```{r}
implied <- vapply(c(0.5, 1, 2, 3), function(ps) {
  x <- leaf_model()
  set_drivers(x, psi_soil = ps, PPFD = 1500)
  x$optimise_psi_stem_TF()
  x$lambda
}, numeric(1))
cat(sprintf(paste0("implied lambda over psi_soil 0.5-3 MPa: ",
                   "%.3g to %.3g umol C / kg, a %.1fx range\n"),
            min(implied), max(implied), max(implied) / min(implied)))
```

So a Cowan-Farquhar λ has to be chosen on the leaf's own scale. A value far below
that band makes water free and pins the optimum wide open; far above it and the
stomata pin shut. Either way the model is not being exercised.

And because both optima satisfy `dA/dE = λ`, handing Cowan-Farquhar the λ that
TF24 *implies* at its own optimum must land it on that same optimum — a check that
needs no derivative approximation anywhere in it:

```{r}
ref <- leaf_model()
set_drivers(ref, psi_soil = 1.0, PPFD = 1500)
ref$optimise_psi_stem_TF()

cf <- leaf_model()
set_drivers(cf, psi_soil = 1.0, PPFD = 1500)
cf$lambda_ <- ref$lambda
cf$optimise_psi_stem_CowanFarquhar()

cat(sprintf(paste0("TF24 optimum %.6f MPa; Cowan-Farquhar at its implied ",
                   "lambda %.6f MPa (gap %.2e)\n"),
            ref$opt_psi_stem_, cf$opt_psi_stem_,
            abs(cf$opt_psi_stem_ - ref$opt_psi_stem_)))
```

## The two gates

Both default off, and both change more than their name suggests.

**Energy balance** (`use_energy_balance_`) makes leaf temperature a function of
transpiration rather than a driver. With it off, a change to any temperature
response is inert at the reference temperature by construction. With it on, `Tleaf`
feeds `vcmax`, `jmax`, `R_d` and the leaf-to-air deficit, and every derivative
picks up a `dT/dE` term.

**Thermal cost** (`use_thermal_cost_`) adds a sigmoid penalty in leaf temperature
— and *also* scales `Jmax` by `(1 − TC)`. So it moves the **benefit** as well as
the cost, which makes it the only gate that changes both sides. It is also inert
without the energy balance, because with a prescribed leaf temperature the penalty
is an additive constant in ψ: it changes the value of the objective and not its
argmax.

## The variants, under one roof

Four axes, chosen independently. A "model" in the literature is a point in this
space, not a separate program.

| axis | options | what it changes |
|---|---|---|
| **cost curve** | TF24 / Cowan-Farquhar / ProfitMax | `C(ψ)`, and its units |
| **supply topology** | single soil potential / multi-layer root network | the decision variable, and the feasible bracket |
| **leaf temperature** | prescribed / energy balance | whether `Tleaf` responds to `E`, and so whether temperature responses are active at all |
| **thermal cost** | off / on | adds a temperature penalty **and** scales `Jmax` — the only axis that moves both sides |

So the published models map on as follows:

- **Cowan & Farquhar** [@cowan1977] — the `λE` cost, prescribed λ. Any topology.
- **Sperry et al.** [@sperry2017] — ProfitMax cost with the normalisation, λ
  derived per operating point. Their own formulation is single-supply.
- **Towers et al.** [@towers2024] ("TF24") — the TF24 carbon cost, solved over the
  collar potential with a multi-layer root network. This is the production path
  here, and the one `plant` consumes.
- **Sicangco et al.** [@sicangco2026] — ProfitMax with the thermal cost on.
- **Jones et al.** [@jones2026] — a multiplicative objective; see the note below.

⚠️ **Jones et al. is not yet in this list as an implemented cost curve.** Its
objective is a **product**, `A · (1 − ψ_leaf/ψ_crit)`, not a difference. Taking
logs turns it into one — `log J = log A + log(1 − ψ/ψ_crit)` — which is the same
additive skeleton with a `log` benefit link, and the argmax is unchanged because
`log` is monotone. Two constraints come with it: `log A` requires `A > 0`, which
fails exactly in the closed-stomata state where `A = −R_d`; and the cost diverges
at `ψ_crit`, so a Jones-type model can never be pinned dry. Both have to be
enforced rather than allowed to become `NaN`.

## The biological components

The optimality argument above is indifferent to what `A` and `E` are. What makes
this a plant model rather than a calculus exercise is the four blocks underneath.

**Photosynthetic demand.** Farquhar-type colimited assimilation [@farquhar1980]: a
Rubisco-limited rate from `Vcmax`, an electron-transport-limited rate from `Jmax`
and incident light, smoothly colimited, minus dark respiration `R_d`. All three
capacities are defined at 25 °C and moved by temperature-response functions, which
is why a grid at 25 °C is blind to every one of them by construction.

**Diffusive supply.** CO₂ enters and water leaves through the same stomata, so one
conductance sets both, related by the ratio of diffusivities. Fixing `E` therefore
fixes `g_c`, hence `c_i`, hence `A` — the chain in "One problem, four coordinates".

**Hydraulic supply.** Water reaches the leaf through a path whose conductivity
*falls* as tension rises, described by the Weibull vulnerability curve of
Appendix A. Transpiration at a given stem potential is the integral of that curve
between the upstream potential and the stem, which is why `E` saturates: pulling
harder recruits progressively less conductance. This saturation is what makes the
optimisation non-trivial — without it, more tension would always buy more carbon.

**The cost of that tension.** The three curves in the table above are three
answers to "what does the plant lose by pulling harder": a carbon penalty
increasing with the fraction of conductivity lost (TF24), the water itself priced
at a constant marginal rate (Cowan-Farquhar), or the normalised loss of remaining
conductance (Sperry). They are not refinements of one another — they disagree about
what is being spent.

**And optionally, energy balance and thermal damage.** A transpiring leaf is
cooler than the air, so `E` feeds back on temperature and hence on every capacity
above. Add a temperature-dependent damage term and the leaf has a reason to
transpire that is not photosynthetic at all.

### Further reading

@sabot2022 compare several of these models in one notation, and are the closest
published analogue to the framing used here. @medlyn2011 give the empirical
conductance model this package can report an equivalent slope for. Full citations
are in the reference list at the end.

## When the leaf cannot transpire

Where the soil is drier than the stem's critical potential there is no flow, and
the operating point is the no-flow one. The answer is to **evaluate the objective
there** rather than to zero the reported fields, so that every output describes one
consistent state. For Cowan-Farquhar this is exactly predictable: `E` is zero, so
the cost is zero, so profit is `−R_d` whatever λ is.

```{r}
d <- leaf_model()                       # psi_soil drier than psi_crit
set_drivers(d, psi_soil = 6.0, PPFD = 1500)
d$lambda_ <- 1.5e5
d$optimise_psi_stem_CowanFarquhar()
cat(sprintf("E = %.1e, cost = %.1e, profit = %.6f, -R_d = %.6f\n",
            d$transpiration_, d$hydraulic_cost_, d$profit_, -d$R_d_))
```

---

# Appendix A — the (P50, c) parameterisation

The Weibull vulnerability curve is conventionally written `exp(-(ψ/b)^c)`. Substituting
`b = P50 / (ln 2)^(1/c)` gives the algebraically identical

$$f(\psi)=2^{-(\psi/P_{50})^{c}}$$

in which **scale and shape are cleanly separated**: `P50` is the potential at 50%
loss, in MPa, and `c` is the steepness. `b` remains available as a derived
read-only quantity, but it is not a parameter anyone should set.

⚠️ **Sign convention.** Published `P50` values are negative (−3.95 MPa). This
package uses positive magnitudes throughout, so `P50` here means `|P50|`.

## Why not `(P50, b)`

`b` *is* a scale parameter — it is the 63% loss point, in the same units and the
same role as `P50`. Specifying both means specifying two points on one curve and
recovering the shape from their ratio:

$$c=\frac{\ln(\ln 2)}{\ln(P_{50}/b)}$$

which diverges as `P50 → b` and is undefined at `P50 = b`, so small errors in
either input make the shape explode. It is also a badly correlated estimation
pair — which is why `fitplc` [@duursma2017] fits `Px` and the *slope* at `Px`
rather than `b` and `c`. `c` is the only genuinely independent second axis.

## Why `c` is not a constant

It carries real biology: `c ≈ 1` is exponential, conductivity bleeding away from
`ψ = 0`; `c > 2` is sigmoid, a safe plateau then rapid failure. Conifers are much
steeper than angiosperms, and leaves and roots are flatter than stems in the same
plant.

Fitted directly in `(P50, c)` from raw PLC curves for eight species:

| species | source / method | \|P50\| (MPa) | c | S50 (%/MPa) |
|---|---|---|---|---|
| *Eucalyptus sideroxylon* | Blackman 2019, bench | 3.95 | 1.99 | 17 |
| *Eucalyptus tereticornis* | Blackman 2019, bench | 4.31 | 2.59 | 21 |
| *Casuarina cunninghamiana* | Blackman 2019, bench | 2.91 | 1.81 | 22 |
| *Pinus radiata* | Blackman 2019, cavitron | 4.09 | 13.0 | 110 |
| *Gossia bidwillii* | Choat 2003 | 5.25 | 0.52 | 3 |
| *Brachychiton australis* | Choat 2003 | 3.49 | 1.21 | 12 |
| *Alphitonia excelsa* | Choat 2003 | 5.69 | 1.70 | 10 |
| *Cochlospermum gillivraei* | Choat 2003 | 1.42 | 5.15 | 126 |

⚠️ **The four Choat 2003 curves are not trustworthy for `c`.** They are 13–14
points each and already sit at 25–30% PLC by 0.8 MPa, which is a native-embolism
or open-vessel artefact rather than a trait. The defensible comparison is the
three Blackman angiosperms measured the same way: **`c` = 1.8–2.6 while `|P50|`
spans 2.9–4.3 MPa** — shape roughly conserved, scale free. That is the evidence
for treating `c` as a parameter with a stable default rather than a constant.

Default `c = 2.0` for angiosperm stems. Conifers need substantially higher; the
13.0 fitted above is inflated by the cavitron technique, and ~4–5 is the safer
literature-style default. Guard `c > 0`, and warn below ~1 — that means no safe
plateau at all, and in practice usually flags a measurement artefact.

## Converting from published numbers

Curves arrive as almost anything except `(P50, c)`. Everything inverts in closed
form. For a remaining conductivity fraction `f` (so `PLC = 1 − f`):

$$P_f = P_{50}\big(\log_2(1/f)\big)^{1/c}
\qquad\Longleftrightarrow\qquad
P_{50} = P_f\big(\log_2(1/f)\big)^{-1/c}$$

From **any two** quantiles, both parameters follow:

$$c=\frac{\ln(L_2/L_1)}{\ln(P_{f_2}/P_{f_1})},\qquad L_i=\log_2(1/f_i)$$

And from `P50` plus the slope at `P50` in % MPa⁻¹, the other commonly published
form:

$$S_{50}=\frac{50\,c\ln 2}{|P_{50}|}=\frac{34.657\,c}{|P_{50}|}$$

Do **not** treat `S50` as the internal parameter — it has `P50` built into it by
construction.

```{r}
L <- function(f) log2(1 / f)                    # f = fraction REMAINING
p_at <- function(P50, c, f) P50 * L(f)^(1 / c)  # any quantile from (P50, c)
c_from_two <- function(P1, f1, P2, f2) log(L(f2) / L(f1)) / log(P2 / P1)
s50_from <- function(P50, c) 50 * c * log(2) / P50

# round-trip: the package defaults, recovered from a (P50, P88) pair
tr <- leaf_traits()
P50 <- tr$stem_b * log(2)^(1 / tr$stem_c)
P88 <- p_at(P50, tr$stem_c, 0.12)
c_hat <- c_from_two(P50, 0.50, P88, 0.12)
P50_hat <- P88 * L(0.12)^(-1 / c_hat)
cat(sprintf("defaults: P50 = %.6f MPa, c = %.6f, so P88 = %.6f, P95 = %.6f\n",
            P50, tr$stem_c, P88, p_at(P50, tr$stem_c, 0.05)))
cat(sprintf("recovered from (P50, P88): c = %.6f, P50 = %.6f  (errors %.1e, %.1e)\n",
            c_hat, P50_hat, abs(c_hat - tr$stem_c), abs(P50_hat - P50)))
cat(sprintf("psi_crit is P95: %.6f vs the trait %.6f\n",
            p_at(P50, tr$stem_c, 0.05), tr$psi_crit))
cat(sprintf("S50 at the defaults: %.1f %%/MPa\n", s50_from(P50, tr$stem_c)))
```

## `psi_crit` is a quantile, not a parameter

`psi_crit` is `P_f` at `f = 0.05`, i.e. `P95` — and **Sperry's reference
conductivity `k_crit = f·kmax` uses the same `f`**. So `psi_crit`, `P95` and
`k_crit` are one number wearing three names. Derived from a single named constant
they cannot disagree; held as a free trait plus a hard-coded `0.05`, they need a
consistency check to police them. `f = 0.05` is a convention, not a knob — it is
0.05 everywhere it appears across this family.

## Two consequences for gradients, and a third

1. **The `stem_b` homogeneity shortcut survives.** At fixed `c`, `b` is a constant
   multiple of `P50`, so the cumulative-vulnerability integral stays homogeneous of
   degree 1 and `d/dP50 = (ln 2)^{-1/c} · d/db`. Reuse the existing fast path with
   a scalar factor rather than writing a new one.
2. **The `c` gradient splits.** Once `P50` is the scale parameter, `b` depends on
   `c` too:
   $$\frac{d}{dc}\bigg|_{P_{50}}=\frac{d}{dc}\bigg|_{b}+\frac{db}{dc}\frac{d}{db},
   \qquad \frac{db}{dc}=\frac{b\ln(\ln 2)}{c^{2}}$$
   The second term is nearly free — it reuses the cheap `b`-derivative — but
   omitting it silently converts a total derivative into a partial, which looks
   like a plausible-but-wrong gradient rather than an error.
3. ⚠️ **A derived `psi_crit` adds a term.** Moving `P50` moves the threshold, so
   `d/dP50` picks up `dpsi_crit/dP50 · d/dpsi_crit`, with `dpsi_crit/dP50 =
   (log₂(1/f))^{1/c}`, a constant at fixed `c`. That term is **exactly zero at an
   interior optimum and order 1 at a pinned one**, so the clean identity in (1)
   holds *only where the optimum is interior*. Assert it with that qualifier.

The root curve obeys the identical homogeneity identity, so it takes the same
treatment — `(root_P50, root_c)` — or the two curves end up on different
conventions.

# Appendix B — gradients

*Placeholder: the implicit-function-theorem composite, `dψ*/dθ = −M/H`, the
envelope theorem and its stationarity condition, and the prescribed-versus-optimum
distinction.*

# Appendix C — numerical practice

*Placeholder: why the argmax is the badly-conditioned quantity and the maximum the
well-conditioned one, the sqrt amplification, the finite-difference step and its
`tol/h` noise floor, and why a scan argmax cannot feed a trait gradient.*
# References
